### 2012-07-31 paper/main.tex
Imported sources from Peter

% 16pages; IFL 2012
%
\documentclass{llncs}
%
\usepackage{amssymb}
\usepackage{cite}
% \usepackage[T1]{fontenc}
% \usepackage[utf8]{inputenc}

%\usepackage{amsmath}
%\usepackage{amsthm}
%\usepackage{proof}
\usepackage{alltt}

\begin{document}
%
\title{Agda Meets Accelerate}
\subtitle{Extended Abstract}
\author{Manuel M T Chakravarty\inst{1} \and Peter Thiemann\inst{2}}
\institute{
University of New South Wales, Sydney, Australia,\\
\email{chak@cse.unsw.edu.au}
\and
  University of Freiburg, Germany,\\
\email{thiemann@informatik.uni-freiburg.de}}

\maketitle              % typeset the title of the contribution

\begin{abstract}
  Are dependent types good for programming in a data parallel
  language?
  We describe the design and implementation of an Agda frontend for
  Accelerate, an embedded DSL for data parallel programming on a
  GPU. We discuss the potential of dependent types in this domain,
  describe some limitations that we ran into, and share some insights
  from our preliminary implementation.
\end{abstract}
\keywords{programming with dependent types, data parallelism}
\thispagestyle{plain}
\pagestyle{plain}
%
\section{Introduction}
\label{sec:introduction}

The future belongs to multi-core computers that require new,
declarative approaches to parallel programming.

Data-parallel programming is one promising candidate for programming
massively parallel machine, in particular the variant that targets
GPUs.

Accelerate \cite{ChakravartyKellerLeeMcdonellGrover2011} is a domain
specific language embedded in Haskell \cite{PeytonJones2003}. Its
target domain is data parallel programming on GPUs.

Effective Parallelism requires many invariants for programs to
run robustly and with predictable performance.

Dependent types are an emerging approach to certified programming,
where invariants are established and proved at compile time.

Thus, it seems fruitful to explore the potential of dependent types in
the context of data-parallel prorgamming.

This paper is a first investigation into this topic. It reports of a
partial port of Accelerate to a new, dependently-typed host language,
Agda \cite{Norell2008,BoveDybjerNorell2009}. Agda is particularly
suited to this port because of its foreign function interface to
Haskell, which enables it to directly invoke functionality of
Accelerate.

Our investigation has the following structure. After recalling some
background on Agda and Accelerate in Section~\ref{sec:background} and
describing related work in Section~\ref{sec:related-work}, 
Section~\ref{sec:dependent-types} discusses potential uses of
dependent types in an array-oriented data parallel language like
Accelerate. Section~\ref{sec:limitations} considers conceptual
problems and limitations that we ran into when constructing the Agda
frontend for Accelerate. Section~\ref{sec:implementation} explains
some technical details of the implementation and discusses some
example code.

\section{Background}
\label{sec:background}

\subsection{Agda}
\label{sec:agda}

Agda \cite{Norell2008,BoveDybjerNorell2009} is a dependently typed
functional programming language. Its basis is a dependently typed
lambda calculus extended with inductive data type families, dependent
records, and parameterized modules. At the same time, Agda is also a
proof assistant for interactively constructing proofs in an
intuitionistic type theory based on the work of Per Martin-L\"of
\cite{MartinLoef1984}.

A familiar example for an indexed data type is the type
\verb+Vec A n+ for vectors of fixed length \verb+n+ and elements of
type \verb+A+ program program with an access operation that restricts the index to the
actual length of the vector at compile time.\footnote{An
  identifier can be an almost arbitrary 
  string of Unicode characters except spaces, parentheses, and curly
  braces. Agda also supports mixfix syntax with the position of
  arguments indicated by underscores in the defining occurrence of an
  identifier.} 
\begin{verbatim}
data Vec (A : Set) : Nat -> Set where
  []   : Vec A zero
  _::_ : {n : Nat} -> A -> Vec A n -> Vec A (suc n)
\end{verbatim}
The above defines an indexed data type \verb+Vec A n+ with two
constructors, \verb+[]+ for the vector of length zero and 
\verb+_::_+ for the infix cons operator that increases the length by one.

One way of writing a safe access operation first defines an indexed
type that encodes the required less-than relation on natural numbers.
\begin{verbatim}
data _<_ : Nat -> Nat -> Set where
  z<s : {n : Nat} -> zero < suc n
  s<s : {m n : Nat} -> m < n -> suc m < suc n
\end{verbatim}
Lines two and three of the definition encode named inference rules for
the cases that $0 < n+1$ (for all $n$) and that $m+1 < n+1$ if $m < n$
(for all $m,n$).

The access operation takes a vector of length \verb+n+, an index
\verb+m+, and a proof of \verb+m < n+ (a derivation tree) to produce
an element of the vector.  
\begin{verbatim}
get : {A : Set} {n : Nat} -> Vec A n -> (m : Nat) -> m < n -> A
get []        _       ()      -- impossible case
get (x :: xs) zero    p       = x
get (x :: xs) (suc m) (s<s p) = get xs m p
\end{verbatim}
This code cannot fail at run time because a caller has to
construct the proof tree for \verb+m < n+ before invoking \verb+get+.

\subsection{Accelerate}
\label{sec:accelerate}

Accelerate \cite{ChakravartyKellerLeeMcdonellGrover2011} is an
embedded DSL for running data parallel code on a GPU.

Accelerate has a rich typed interface that already encodes many
invariants using advanced typing concepts in Haskell, in particular
GADTs\cite{PeytonJonesVytiniotisWeirichWashburn2006}, associated types
\cite{ChakravartyKellerJones2005}, and type functions
\cite{SchrijversPeytonJonesChakravartySulzmann2008}.

Moreover, Accelerate is a \emph{generative library}, in the sense that
most of its API operations just construct abstract syntax trees (AST). In the end,
the \verb+run+ operation accepts such an AST (of type \verb+Acc a+),
compiles it to GPU kernels, 
uploads it to the device, executes it, and retrieves the results. 
\begin{verbatim}
CUDA.run :: Arrays a => Acc a -> a
\end{verbatim}
The type class constraint \verb+Arrays a+ restricts the result type to
a type that can be handled by accelerate, usually an array type.

A simple Accelerate program would look like this:
\begin{verbatim}
dotp :: Vector Float -> Vector Float -> Acc (Scalar Float)
dotp xs ys = let xs' = use xs
                 ys' = use ys
             in  fold (+) 0 (zipWith (*) xs' ys')
\end{verbatim}
Here, \verb+Vector+ and \verb+Scalar+ are one- and zero-dimensional
array types known to Accelerate. Initially, the values of the
\verb+Vector+ type are simply Haskell arrays. The \verb+use+ operation
transforms a Haskell array into an array reference on the GPU and
makes the reference available for use in constructing the AST of the
desired operation. The last line applies two API operations,
\texttt{fold} and \texttt{zipWith} to construct an AST that returns a
\texttt{Scalar Float}, i.e., a single float value. This line relies
heavily on (type class) overloading: \texttt{0}, \texttt{(+)}, and
\texttt{(*)} are overloaded to just construct abstract syntax.

\textbf{TODO}: explain type \texttt{Array sh a}, in particular the
\texttt{sh} part.

\section{Related Work}
\label{sec:related-work}

\section{Dependent Types for Accelerate}
\label{sec:dependent-types}

In this section, we investigate the potential uses of dependent typing
in a language like Accelerate.

\subsection{Exact Checking of Array Bounds}
\label{sec:exact-checking-array}

Accelerate's API features expressive type constraints that describe
the shape of the array arguments and results. These constraints ensure
that no shape mismatches occur (e.g., a one-dimensional array is
considered two-dimensional). However, they do not ensure at compile
time that the sizes of the dimensions match up.

As an example, consider the function \texttt{reshape}. 
It takes a target shape \texttt{sh} and an array of source shape
\texttt{sh'} and changes the layout of that array to \texttt{sh}. 
\begin{verbatim}
reshape :: Exp sh -> Acc (Array sh' e) -> Acc (Array sh e)
\end{verbatim}
For this reshaping to work correctly, the underlying
number of elements must remain the same. For example, while it makes sense
to reshape a two-dimensional $3\times 4$-array  to a vector of size
$12$ or to a three-dimensional $3\times2\times2$-array, whereas an attempt to
reshape the same array to a $2\times5$-array should be rejected at
compile time.

In Agda, we define the shape type as follows.
\begin{verbatim}
data Shape : Set where
  Z     : Shape
  _:<_> : Shape -> Nat -> Shape
\end{verbatim}
The \texttt{size} function computes the number of elements stored in a
shape.
\begin{verbatim}
size : Shape -> Nat
size Z = 1
size (sh :< n >) = size sh * n
\end{verbatim}
Now we can state an accurate type for \texttt{reshape} in Agda, which
involves an extra argument with a proof that the source and
target shapes have the same size.\footnote{For the moment, the type
  \texttt{Element} could be read as \texttt{Set}. Actually,
  \texttt{Element} classifies the types of values that can be stored
  in Accelerate arrays. In Haskell, this restriction is imposed with a
  type class.}
\begin{verbatim}
reshape : {sh : Shape} {E : Element}
       -> (sh' : Shape) -> Array sh E -> (size sh = size sh')
       -> Array sh' E
\end{verbatim}
There is a subtle difference to the original signature. In
Accelerate, the first argument is an \emph{expression} that produces a
value of type \texttt{sh} at run time, whereas the Agda
\texttt{reshape} requires a \texttt{Shape} as its first argument.

Furthermore, functions like \texttt{map} and \texttt{zipWith} obtain
more precise types. The type of \texttt{map} tells us that the input
shape is identical to the output shape:
\begin{verbatim}
map : {A B} {sh} -> (Exp A -> Exp B) -> Array sh A -> Array sh B
\end{verbatim}
Similarly, the type of \texttt{zipWith} restricts its input arrays to
identical shapes:
\begin{verbatim}
zipWith : {A B C} {sh} -> (Exp A -> Exp B -> Exp C)
        -> Array sh A -> Array sh B -> Array sh C
\end{verbatim}
The latter type is more restrictive than the Accelerate
implementation of \texttt{zipWith}. Instead of checking the dimensions
of the input arrays, it truncates them to the respective minima. A
corresponding Agda type could be developed easily. It would involve
specifying a suitable ternary relation between shapes.

\subsection{Associativity of Operations}
\label{sec:assoc-oper}

Some parallel reduction operations require their base operation to be
associative to return a predictable result. Here are two examples from
Accelerate. 
\begin{verbatim}
fold  :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) -> Exp a ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
fold1 :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
\end{verbatim}
In both cases, the text of the documentation says that ``the first
argument needs to be associative'' and the \texttt{fold1}
documentation ``requires the reduced array to be non-empty''.
The second requirement can be enforced by asking for a suitable proof
object on each call of \texttt{fold1}:
\begin{verbatim}
fold1 : ... -> Array (sh :< n >) E -> (p : size sh * n > 0)
            -> Array sh E
\end{verbatim}
The first requirement can be rephrased to saying that the first two
parameters of \texttt{fold} together form a monoid, which requires an
associative operation with a unit element. The concept of a monoid
can be formalized in Agda, which has indeed been done in the standard
library. Unfortunately, the formalization from the library cannot be
used because Accelerate deals with ASTs, not with values. So, a
formalization is required that states that the meaning of an
AST-encoded function is associative and the meaning of another
AST-encoded constant is its unit element. Given that Accelerate
encodes AST construction using higher-order abstract syntax, such a
formalization is not straightforward. Moreover, even given expressions
with a fixed meaning, there is no general shape for associative
functions, so that proofs can only be done for special cases.

\subsection{Embedding of Constants}
\label{sec:embedding-constants}

Accelerate relies on Haskell's type class \texttt{Num} and built-in
magic to embed constants. The magic is that Haskell scans each integer
literal as a value of type \texttt{Integer}, which is a built-in type
of unlimited integers. To this scanned value, Haskell applies the
function \texttt{fromInteger} to convert to the expected
type. Similarly, floating point constants are scanned as values of
type \texttt{Rational} (\texttt{Integer} fractions) and then converted
using \texttt{fromRational}. Accelerate provides suitable instances of
the numeric type classes that define \texttt{fromInteger} and
\texttt{fromRational} to produce suitable AST fragments.

In Agda, the situation is more involved because it only supports
natural number literals which abbreviate a unary encoding. Therefore,
we embed a numeric literal using a string with an explicit
type annotation that determines the parsing of the string. Here are
some example embeddings:
\begin{verbatim}
"3.1415926" ::: Float
"6.0221415E23" ::: Double
\end{verbatim}
It should be noted that \texttt{Float} and \texttt{Double} are not
types, but rather values of type \texttt{Element} as all other Haskell
types, which can be referred to in an Accelerate program:
\begin{verbatim}
data Element : Set where
  Bool : Element
  Int : Element
  Float : Element
  Double : Element
  ...
\end{verbatim}
All magic of the embedding is hidden in the \texttt{:::} operation:
\begin{verbatim}
_:::_ : (s : String) -> (E : Element) 
     -> {{nu : IsNumeric E}} -> {p : T (s parsesAs E)} -> Exp E
s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
\end{verbatim}
The argument \texttt{nu} is an instance argument that is automatically
filled-in with a suitably typed value in scope. Here, the predicate
\texttt{IsNumeric} plays the role of a type class that characterizes
the numeric types.
\begin{alltt}
IsNumeric : Element -> Set
IsNumeric Int = \(\top\)
IsNumeric Float = \(\top\)
IsNumeric Double = \(\top\)
IsNumeric _ = \(\bot\)
\end{alltt}
The function \texttt{parsesAs} dispatches on its ``type'' argument and
checks whether the string is a constant of the expected type. The
function \texttt{constantFromString} is imported from Accelerate.
It is an overloaded function that requires two type dictionaries, thus
they are computed from \texttt{E} using the function \texttt{EltDict}
and \texttt{ReadDict}.

\section{Limitations}
\label{sec:limitations}

In a number of places, Accelerate's generativity limits the
applicability of dependent typing. We already mentioned that the
formalization of associativity or of the concept monoid gets
unmanageable because it has to be asserted for abstract syntax.

For a similar reason, an implementation of the \texttt{filter}
operation creates problems. The problem is that the size of the result
cannot be determined statically.

Existential types do not help with this problem.

However, an alternative encoding dependently-typed of arrays can be
used which is compatible with filtering of elements. Actually, there
are several ways to implement such an encoding with different drawbacks.

\section{Implementation}
\label{sec:implementation}

Agda is compiled to Haskell.

The compiler supports a Haskell foreign function interface.

The sophisticated type structure of the Accelerate language is in the
way.

The foreign function interface is not sufficiently expressive so that
Agda types need to be encoded and corresponding structures need to be
build from this encoding in the Haskell code.

\section{Conclusion}
\label{sec:conclusion}

%
% ---- Bibliography ----
%
\bibliography{abbrevs,papers,books,collections,misc,theses}
\bibliographystyle{abbrv}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2012-07-31 code/Accel.hs

@@ -1,0 +1,401 @@
+{-# LANGUAGE GADTs, TypeOperators #-}
+module Accel
+  (Array,
+   EltDict(..), eltDictZZ, eltDictZC, eltDictBool, eltDictInt, eltDictFloat, eltDictDouble, eltDictPair, 
+   IsNumDict(..), isNumDictInt, isNumDictFloat, isNumDictDouble,
+   ReadDict(..), readDictBool, readDictInt, readDictFloat, readDictDouble,
+   SizedShapeDict(..), sizedShapeDictZ, sizedShapeDictC,
+   ShapeDict(..), shapeDictZ, shapeDictC,
+   IsScalarDict(..), isScalarDictBool, isScalarDictInt, isScalarDictDouble, isScalarDictFloat,
+   IsIntegralDict(..), isIntegralDictInt,
+   BitsDict(..), bitsDictInt,
+   constantInt, constantFloat, constantDouble, constantFromString, valFromString,
+   constant, mkAdd, mkSub, mkMul, mkAbs, mkIte, mkBXor, mkShiftR,
+   fromList, fromList', toList, showArray,
+   use, run, generate, fold, fold1, Accel.scanl, scanl', Accel.scanl1, prescanl, prescanr, permute, backpermute,
+   Accel.zipWith, Accel.zip, Accel.map, the, (!),
+   Accel.fst, Accel.snd, Accel.pair
+  ) where
+
+import qualified Data.Bits as B 
+import qualified Data.Array.Accelerate as A
+import qualified Data.Array.Accelerate.Interpreter as Interpreter
+import qualified Data.Array.Accelerate.Smart as Smart
+import Unsafe.Coerce
+
+-- element dictionaries -- needed for array element types
+data EltDict e where
+  EltDict :: (A.Elt e) => EltDict e
+
+eltDictZZ :: EltDict A.Z
+eltDictZZ = EltDict
+
+eltDictZC :: EltDict sh -> EltDict ix -> EltDict ((A.:.) sh ix)
+eltDictZC EltDict EltDict = EltDict
+
+eltDictPair :: EltDict a -> EltDict b -> EltDict (a, b)
+eltDictPair EltDict EltDict = EltDict
+
+eltDictBool :: EltDict Bool
+eltDictBool = EltDict
+
[…]

### 2012-07-31 code/Generator.hs

@@ -1,0 +1,26 @@
+module Generator where
+
+import qualified Data.Array.Accelerate as A
+import Accel
+import System.Random
+
+fixedArray111 :: IO (Array A.DIM1 Int)
+fixedArray111 = return (fromList' [1, 1, 1])
+
+randomArray52 :: IO (Array A.DIM1 Int)
+randomArray52 = randomArray 52
+
+randomArray :: Int -> IO (Array A.DIM1 Int)
+randomArray n = do
+  xs <- randomList n
+  return (fromList' xs)
+
+randomList :: Int -> IO [Int]
+randomList 0 = 
+  return []
+randomList n = do
+  xs <- randomList (n-1)
+  x <- randomInt
+  return (x : xs)
+
+randomInt = getStdRandom (randomR (minBound,maxBound))

### 2012-07-31 paper/main.tex
Revised the abstract

@@
 \subtitle{Extended Abstract}
-\author{Manuel M T Chakravarty\inst{1} \and Peter Thiemann\inst{2}}
+\author{Peter Thiemann\inst{1} \and Manuel M. T. Chakravarty\inst{2}}
 \institute{
+  University of Freiburg, Germany,\\
+  \email{thiemann@informatik.uni-freiburg.de}
+\and
 University of New South Wales, Sydney, Australia,\\
 \email{chak@cse.unsw.edu.au}
-\and
-  University of Freiburg, Germany,\\
-\email{thiemann@informatik.uni-freiburg.de}}
+}
 
@@
 \begin{abstract}  
-  Are dependent types good for programming in a data parallel
-  language?
-  We describe the design and implementation of an Agda frontend for
-  Accelerate, an embedded DSL for data parallel programming on a
-  GPU. We discuss the potential of dependent types in this domain,
-  describe some limitations that we ran into, and share some insights
-  from our preliminary implementation.
+  Embedded languages in Haskell benefit from a range of type extensions, such as type families, that are subsumed by dependent types.
+However, even with those type extensions, embedded languages for data parallel programming lack desirable static guarantees, such as static bounds checks in indexing and collective permutation operations.
+
+  This raises the question whether embedded languages for data parallel programming would benefit from fully-fledged dependent types, such as those available in Agda.
+We explored that question by designing and implementing an Agda frontend to Accelerate, an embedded language for data parallel programming aimed at GPUs.
+We discuss the potential of dependent types in this domain, describe some of the limitations that we encountered, and share some insights from our preliminary implementation.
 \end{abstract}

### 2012-07-31 paper/main.tex
Started revising the introduction

@@
 
-The future belongs to multi-core computers that require new,
-declarative approaches to parallel programming.
+Generative approaches to programming parallel hardware promise to combine high-level programming models with high-performance.
+They are especially attractive for targeting restricted architectures, such as GPUs (graphics processor units), that cannot efficiently execute code aimed at conventional multicore CPUs.
+Instead, GPUs require a high degree of data parallelism, restricted control flow, and carefully tailored data access patterns to be efficient.
+Previous work ---for example, Accelerator~\cite{??}, Copperhead~\cite{??}, and Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011}--- demonstrates that embedded array languages with a custom code generator can meet those GPU constraints by carefully limiting the embedded language constructs.
 
-Data-parallel programming is one promising candidate for programming
-massively parallel machine, in particular the variant that targets
-GPUs.
-
-Accelerate \cite{ChakravartyKellerLeeMcdonellGrover2011} is a domain
-specific language embedded in Haskell \cite{PeytonJones2003}. Its
-target domain is data parallel programming on GPUs.
-
-Effective Parallelism requires many invariants for programs to
-run robustly and with predictable performance.
+Given a host languages with an expressive type system, it is attractive to leverage that type system to express static properties of the embedded language.
+For example, Accelerate, an embedded array language for Haskell, uses Haskell's recent support for type-level programming like GADTs and type families in that manner~\cite{ChakravartyKellerLeeMcdonellGrover2011}.
+This is especially important for approaches relying on runtime code generation as we want to mimise the number of potential faults at application runtime, which coincides with the compile time of the embedded language.
+Moreover, static guaranties hold the potential to improve the predictability of parallel performance.
 

### 2012-07-31 paper/main.tex
More in the introduction

@@
 
-Dependent types are an emerging approach to certified programming,
-where invariants are established and proved at compile time.
-
-Thus, it seems fruitful to explore the potential of dependent types in
-the context of data-parallel prorgamming.
+Dependent types are an emerging approach to certified programming, where invariants are established in the form of types and proved at compile time.
+Many of Haskell's type-level extensions used in Accelerate approximate various aspects of dependently-typed programming.
+Hence, it it is natural to ask whether fully-fledged dependent types, such as those provided by Agda, improve the specification of an embedded language like Accelerate, whether they increase the scope of static guarantees, and whether they may be leveraged to predict performance more accurately.
 

### 2012-07-31 paper/main.tex
Revised and extended the Accelerate subsection

@@
 
-Accelerate \cite{ChakravartyKellerLeeMcdonellGrover2011} is an
-embedded DSL for running data parallel code on a GPU.
-
-Accelerate has a rich typed interface that already encodes many
-invariants using advanced typing concepts in Haskell, in particular
-GADTs\cite{PeytonJonesVytiniotisWeirichWashburn2006}, associated types
-\cite{ChakravartyKellerJones2005}, and type functions
-\cite{SchrijversPeytonJonesChakravartySulzmann2008}.
-
-Moreover, Accelerate is a \emph{generative library}, in the sense that
-most of its API operations just construct abstract syntax trees (AST). In the end,
-the \verb+run+ operation accepts such an AST (of type \verb+Acc a+),
-compiles it to GPU kernels, 
-uploads it to the device, executes it, and retrieves the results. 
+Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011} is a data-parallel array language embedded into Haskell, which targets GPUs.
+It is a \emph{generative library}, as its data-parallel array operations are not executed directly, but instead construct abstract syntax trees (AST) representing an entire data-parallel subcomputation.
+These \emph{computation representations} are executed using a \verb+run+ operation that accepts such an AST (of type \verb+Acc a+), compiles it to GPU kernels, uploads it to the device, executes it, and retrieves the results.
+%
 \begin{verbatim}
@@
 \end{verbatim}
-The type class constraint \verb+Arrays a+ restricts the result type to
-a type that can be handled by accelerate, usually an array type.
+%
+The type class constraint \verb+Arrays a+ restricts the result type to a single array or a tuple of arrays.
 
-A simple Accelerate program would look like this:
+As computation representations of type \verb+Acc a+ are compiled at application runtime, all \verb+Acc+ compilation errors are effectivly \emph{runtime errors} of the application.
+Hence, Accelerate uses a range of Haskell type system extensions to statically type Accelerate expressions, such that these runtime errors are avoided where possible.
+In particular, Accelerate uses GADTs\cite{PeytonJonesVytiniotisWeirichWashburn2006}, associated types
+\cite{ChakravartyKellerJones2005}, and type families
+\cite{SchrijversPeytonJonesChakravartySulzmann2008}.
+
+As a simple example of an Accelerate program, consider a function implementing a dot product:
+%
 \begin{verbatim}
@@
 \end{verbatim}
-Here, \verb+Vector+ and \verb+Scalar+ are one- and zero-dimensional
-array types known to Accelerate. Initially, the values of the
-\verb+Vector+ type are simply Haskell arrays. The \verb+use+ operation
-transforms a Haskell array into an array reference on the GPU and
-makes the reference available for use in constructing the AST of the
-desired operation. The last line applies two API operations,
-\texttt{fold} and \texttt{zipWith} to construct an AST that returns a
-\texttt{Scalar Float}, i.e., a single float value. This line relies
+%
+The types \verb+Vector+ and \verb+Scalar+ represent one- and zero-dimensional
+arrays.
+Plain arrays, such as \verb+Vector Float+ are conventional Haskell arrays, using an unboxed representation to improve performance.
+However, when they are wrapped into the constructor \verb+Acc+, such as in \verb+Acc (Scalar Float)+, they represent arrays of the embedded language and are allocated in GPU memory, which in current high-performance GPUs is physically separate from CPU memory.
+
+The \verb+use+ operation makes a Haskell array available in the embedded language by wrapping it into the \verb+Acc+ constructor and copying it to GPU memory.\footnote{Accelerate employs a caching strategy to avoid the transfer of arrays, which are already available in GPU memory.} The operations \texttt{fold} and \texttt{zipWith} represent collective operations on Accelerate arrays, effectively producing a representation of an array computation yielding a value of
+\texttt{Scalar Float}; i.e., a single float value. This code relies
 heavily on (type class) overloading: \texttt{0}, \texttt{(+)}, and
@@
 
-\textbf{TODO}: explain type \texttt{Array sh a}, in particular the
-\texttt{sh} part.
+The type \verb+Scalar+ and \verb+Vector+ are merely type synonyms instantiating a shape-parameterised array type to the special case of zero and one dimensional arrays:
+%
+\begin{verbatim}
+type Scalar e = Array DIM0 e
+type Vector e = Array DIM1 e
+\end{verbatim}
+%
+So, the general type for \verb+use+ is
+%
+\begin{verbatim}
+use :: Elt e => Array sh e -> Acc (Array sh e)
+\end{verbatim}
+%
+where the class \verb+Elt+ characterises all types that may be held in Accelerate arrays. These are currently primitive types and tuples.
+
+Common dimensions, such as \verb+DIM0+, \verb+DIM1+, and so on, are predefined, but to enable shape polymorphic computations, along the lines pioneered in the Haskell array library Repa~\cite{keller-etal:repa}, shapes are inductively defined using type-level snoc lists formed from
+%
+\begin{verbatim}
+data Z       = Z
+data sh :. i = sh :. i
+\end{verbatim}
+%
+Hence, we have got
+%
+\begin{verbatim}
+type DIM0 = Z
+type DIM1 = DIM0 :. Int
+<and so on>
+\end{verbatim}
 

### 2012-07-31 paper/main.tex
Added some related work
I inlined the citations as I haven't got the bib files at the moment.

@@
 \label{sec:related-work}
+
+Dependent types for array programming: 
+\begin{itemize}
+\item "Dependent Types for Distributed Arrays".
+Wouter Swierstra and Thorsten Altenkirch. Postproceedings of TFP 2008.
+\item "More Dependent Types for Distributed Arrays".
+Wouter Swierstra. Accepted for publication in the Journal of Higher-order and Symbolic Computation.
+\end{itemize}
+This work focuses on describing distribution strategies using dependent types (which is orthogonal to our work). However, to characterise distributions, it also tracks bounds in types.
+
+DML:
+\begin{itemize}
+\item Xi, Hongwei (March 2007).
+"Dependent ML: An Approach to Practical Programming with Dependent Types".
+Journal of Functional Programming 17 (2).
+[There is a previous ICFP paper co-authored with his advisor Bob Harper.]
+\end{itemize}
+Extension of ML to track array bounds.
+
+Other embedded array languages:
+\begin{itemize}
+\item Accelerator: C++ library providing data-parallel GPU programming; subsequent work by Satnam Singh provides an F\# binding (but that was never published as for as I know---double check); the language provides only a small subset of the functionality of Accelerate
+\item Copperhead: embedded array language in Python; no static guarantees; performance varied widely in our experiments
+\item anything else?
+\end{itemize}
 

### 2012-07-31 paper/main.tex
section 4 and 5

@@
 
-\subsection{Embedding of Constants}
+\subsection{Embedding of Haskell Types}
 \label{sec:embedding-constants}
 
-Accelerate relies on Haskell's type class \texttt{Num} and built-in
-magic to embed constants. The magic is that Haskell scans each integer
-literal as a value of type \texttt{Integer}, which is a built-in type
-of unlimited integers. To this scanned value, Haskell applies the
-function \texttt{fromInteger} to convert to the expected
-type. Similarly, floating point constants are scanned as values of
-type \texttt{Rational} (\texttt{Integer} fractions) and then converted
-using \texttt{fromRational}. Accelerate provides suitable instances of
-the numeric type classes that define \texttt{fromInteger} and
-\texttt{fromRational} to produce suitable AST fragments.
-
-In Agda, the situation is more involved because it only supports
-natural number literals which abbreviate a unary encoding. Therefore,
-we embed a numeric literal using a string with an explicit
-type annotation that determines the parsing of the string. Here are
-some example embeddings:
-\begin{verbatim}
-"3.1415926" ::: Float
-"6.0221415E23" ::: Double
-\end{verbatim}
-It should be noted that \texttt{Float} and \texttt{Double} are not
-types, but rather values of type \texttt{Element} as all other Haskell
-types, which can be referred to in an Accelerate program:
+Accelerate supports a wide range of numeric types as base types for
+array computations. Almost all of these types have no suitable
+counterpart in Agda, which only supports natural numbers in unary
+encoding. For that reason, our embedding keeps these types 
+abstract in Agda. To specify functions that are polymorphic in such a
+Haskell type or depend on it in some way, we have reified these types
+in an Agda type \texttt{Element}.
 \begin{verbatim}
@@
   Double : Element
+  Pair   : Element -> Element -> Element
   ...
 \end{verbatim}
-All magic of the embedding is hidden in the \texttt{:::} operation:
-\begin{verbatim}
-_:::_ : (s : String) -> (E : Element) 
-     -> {{nu : IsNumeric E}} -> {p : T (s parsesAs E)} -> Exp E
-s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
-\end{verbatim}
-The argument \texttt{nu} is an instance argument that is automatically
-filled-in with a suitably typed value in scope. Here, the predicate
-\texttt{IsNumeric} plays the role of a type class that characterizes
-the numeric types.
+Corresponding to Haskell type classes that are used in Accelerate, our
+embedding supplies predicates that characterize subsets. For example,
+the set of numeric type is defined by a predicate \texttt{IsNumeric}.
 \begin{alltt}
@@
 \end{alltt}
+
+\subsection{Embedding of Constants}
+\label{sec:embedding-constants}
+
+Accelerate relies on Haskell's built-in support for the type classes
+\texttt{Num} and \texttt{Fractional} to embed constants. The Haskell compiler reads each integer
+literal as a value of type \texttt{Integer}, which is a built-in type
+of arbitrary precision integers. To this value, Haskell applies the
+function \texttt{fromInteger} that converts to the expected
+type. Similarly, floating point constants are read as values of
+type \texttt{Rational} (\texttt{Integer} fractions) and then converted
+using \texttt{fromRational}. Accelerate provides instances of
+these type classes that define \texttt{fromInteger} and
+\texttt{fromRational} to produce suitable AST fragments.
+
+Because of Agda's limited support for numeric data types,
+we embed more ambitious numeric literals for floating point numbers
+using a string with an explicit type annotation that determines the
+parsing of the string. Here are some example embeddings:
+\begin{verbatim}
+"3.1415926" ::: Float
+"6.0221415E23" ::: Double
+\end{verbatim}
+Recall that \texttt{Float} and \texttt{Double} are not
+types, but rather values of type \texttt{Element}.
+All magic of the embedding is hidden in the \texttt{:::} operation:
+\begin{verbatim}
+_:::_ : (s : String) -> (E : Element) 
+     -> {{nu : IsNumeric E}} -> {p : T (s parsesAs E)} -> Exp E
+s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
+\end{verbatim}
+The arguments \texttt{s} and \texttt{E} are explicit, but the
+remaining ones are to be inferred by Agda.
+The argument \texttt{nu} is an instance argument that is automatically
+filled-in with a suitably typed value in scope
+\cite{DevriesePiessens2011}. Here, the predicate 
+\texttt{IsNumeric} plays the role of a type class that characterizes
+the numeric types.
+
 The function \texttt{parsesAs} dispatches on its ``type'' argument and
@@
 function \texttt{constantFromString} is imported from Accelerate.
-It is an overloaded function that requires two type dictionaries, thus
-they are computed from \texttt{E} using the function \texttt{EltDict}
+It is an overloaded function that requires two type dictionaries,
+which are computed from \texttt{E} using the functions \texttt{EltDict}
 and \texttt{ReadDict}.
@@
 formalization of associativity or of the concept monoid gets
-unmanageable because it has to be asserted for abstract syntax.
+unmanageable because such properties have to be asserted for abstract syntax.
 
-For a similar reason, an implementation of the \texttt{filter}
-operation creates problems. The problem is that the size of the result
-cannot be determined statically.
+For a similar problem, consider an implementation of the \texttt{filter}
+operation that takes a predicate and a source array and returns an
+array that only contains the elements of the source array fulfilling
+the predicate.  First of all, filtering only makes sense for
+one-dimensional arrays, that is, for vectors. To see the second catch,
+let's try to write down a dependent type signature for \texttt{filter}.
+\begin{verbatim}
+filter : {n m : Nat}{E : Element}
+       -> Vector n E -> (Exp E -> Exp Bool) -> Vector m E
+\end{verbatim}
+The problem is that the size of the result
+cannot be determined statically. In fact, the only thing we know about
+\texttt{m} is that it must be less than or equal to
+\texttt{n}. However, we cannot prove this from the code because the
+Accelerate implementation computes the length of the result in a quite
+subtle way. Moreover, as Accelerate deals with the construction of
+abstract syntax most of the time, the value of \texttt{m} is simply
+not available at run time of \texttt{filter}.
 
-Existential types do not help with this problem.
+One might contemplate that an existential type like
+\begin{verbatim}
+exists (\ m : Nat -> m <= n -> Vector m E)
+\end{verbatim}
+could be employed. However, it is not possible to build such an
+existential package because the evidence \texttt{m} is not available
+when the package is constructed.
 
-However, an alternative encoding dependently-typed of arrays can be
-used which is compatible with filtering of elements. Actually, there
-are several ways to implement such an encoding with different drawbacks.
+However, an alternative encoding of arrays can be
+used which is compatible with filtering of elements. The idea of this
+encoding is to keep all elements but mark those which are no longer
+present because they have been filtered out.
+There are several ways of implementing this idea. The simplest
+approach is to wrap each element in a maybe type.
+pair up each element with a boolean flag that indicates
+its presence.\footnote{A \texttt{Maybe} type would be a better option,
+  but it has to be coded without using pattern matching.}
+\begin{verbatim}
+FVector : Nat -> Element -> Set
+FVector n E = Vector n (Pair Bool E)
+\end{verbatim}
+
+Now filtering becomes quite simple.
+\begin{verbatim}
+filterF : {n : Nat}{E : Element}
+        -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
+filterF {n}{E} pred vec =
+  map g vec
+  where
+  g : Exp (Pair Bool E) -> Exp (Pair Bool E)
+  g bx = pair ((fst bx) && p (snd bx)) x
+\end{verbatim}
+However, mapping becomes more complicated because it either has to
+materialize a dummy result for each absent element in the argument
+vector or apply the function to absent elements, too.\footnote{This
+  complication could be avoided with the \texttt{Maybe} type.}
+\begin{verbatim}
+mapF : {n : Nat}{E F : Element}
+     -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
+mapF {n}{E}{F} defaultF f vec =
+  map g vec
+  where
+  g : Exp (Pair Bool E) -> Exp (Pair Bool F)
+  g bx = if (fst bx) then (pair (fst bx) (f (snd bx)))
+                     else (pair (fst bx) defaultF)
+\end{verbatim}
+On the positive side, some operations can get rid of the absent
+elements. In particular, a fold operation which reduces a filtered
+vector with a monoid returns a single value. In Accelerate, such a value has type
+\texttt{Scalar}, which is a synonym for an array of dimension $0$.
+\begin{verbatim}
+foldF : {n : Nat}{E : Element}
+      -> (Exp E -> Exp E -> Exp E) -> Exp E
+      -> FVector n E -> Scalar E
+foldF f e vec =
+  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
+\end{verbatim}
+Other operations like \texttt{fold1} and the scan operations present
+in Accelerate can also be lifted to this representation, but they
+retain a notion of absent elements and do not allow to revert to a
+non-filtered representation.
+
+In the end, such a representation may not be a loss on a GPU. As long
+as all computations take the same path, all processing elements work
+in unison. As soon as there are different paths in the same
+computation step, then some elements will be idle for part of the
+computation step. So it would be most advantageous to organize work as
+uniformly as possible.
 

### 2012-08-01 paper/main.tex
Revise related work

@@
 \begin{verbatim}
+data Nat : Set where
+  zero : Nat
+  suc  : Nat -> Nat
+
 data Vec (A : Set) : Nat -> Set where
@@
 
-Dependent types for array programming: 
-\begin{itemize}
-\item "Dependent Types for Distributed Arrays".
-Wouter Swierstra and Thorsten Altenkirch. Postproceedings of TFP 2008.
-\item "More Dependent Types for Distributed Arrays".
-Wouter Swierstra. Accepted for publication in the Journal of Higher-order and Symbolic Computation.
-\end{itemize}
-This work focuses on describing distribution strategies using dependent types (which is orthogonal to our work). However, to characterise distributions, it also tracks bounds in types.
+Peebles formalised parts of the API of the Haskell array library Repa using Agda~\cite{peebles:derpa}.
+The formalisation uses the same shape structure as we are using in Accelerate, but array computations are neither embedded nor can parallel high-performance code be generated.
 
-DML:
-\begin{itemize}
-\item Xi, Hongwei (March 2007).
-"Dependent ML: An Approach to Practical Programming with Dependent Types".
-Journal of Functional Programming 17 (2).
-[There is a previous ICFP paper co-authored with his advisor Bob Harper.]
-\end{itemize}
-Extension of ML to track array bounds.
+Swierstra and Altenkirch investigated the use of dependent types for \emph{distributed} array programming~\cite{swierstra-altenkirch:dep-types-for-distr-arrays,swierstra:more-dependent-types}.
+Their notation for distributed arrays was inspired by X10 and the main focus is on expressing locality aware algorithms.
 
-Other embedded array languages:
-\begin{itemize}
-\item Accelerator: C++ library providing data-parallel GPU programming; subsequent work by Satnam Singh provides an F\# binding (but that was never published as for as I know---double check); the language provides only a small subset of the functionality of Accelerate
-\item Copperhead: embedded array language in Python; no static guarantees; performance varied widely in our experiments
-\item anything else?
-\end{itemize}
+Dependent ML is an ML dialect with restricted form of dependent types, which, among other applications, may be used to statically check array bounds in array operations~\cite{xi:dml-jfp}.
+However, only simple indexing and array updating are considered and not aggregate array operations, such as those provided by Accelerate.
+
+Accelerator~\cite{Tarditi:2006} enables embedded GPU computations in C\# programs; it subsequently also added F\# support.
+However, no attempt is made to track properties of array programs statically.
+Similarly, Copperhead~\cite{Catanzaro:EECS-2010-124} embedded an array language into Python, but does not attempt to track information statically.
 
@@
 %
-\bibliography{abbrevs,papers,books,collections,misc,theses}
+\bibliography{local,abbrevs,papers,books,collections,misc,theses}
 \bibliographystyle{abbrv}

### 2012-08-01 paper/main.tex
section 3 4 5

--- 2012-07-31 paper/main.tex
@@
 \usepackage{cite}
+\usepackage{color}
 % \usepackage[T1]{fontenc}
-% \usepackage[utf8]{inputenc}
 
@@
 \usepackage{alltt}
+
+% UNICODE in Agda sources: more trouble than it is worth right now!
+
+% \usepackage{bbm}
+% \usepackage[greek,english]{babel}
+% \usepackage{ucs}
+% \usepackage[utf8x]{inputenc}
+% %\usepackage{autofe}
+% \DeclareUnicodeCharacter{8704}{\ensuremath{\forall}}
+% % \DeclareUnicodeCharacter{8594}{\ensuremath{\mathbb{N}}}
+% \DeclareUnicodeCharacter{8594}{\ensuremath{\to}}
+
+\usepackage{fancyvrb}
+
+\DefineVerbatimEnvironment{code}{Verbatim}{} % Add fancy options here if you like.
+\DefineVerbatimEnvironment{hcode}{Verbatim}{formatcom=\color{blue},frame=leftline} % Add fancy options here if you like.
 
@@
 
-Agda \cite{Norell2008,BoveDybjerNorell2009} is a dependently typed
-functional programming language. Its basis is a dependently typed
+Agda \cite{Norell2008,BoveDybjerNorell2009} is a dependently-typed
+functional programming language. Its basis is a dependently-typed
 lambda calculus extended with inductive data type families, dependent
@@
 
-Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011} is a data-parallel array language embedded into Haskell, which targets GPUs.
-It is a \emph{generative library}, as its data-parallel array operations are not executed directly, but instead construct abstract syntax trees (AST) representing an entire data-parallel subcomputation.
-These \emph{computation representations} are executed using a \verb+run+ operation that accepts such an AST (of type \verb+Acc a+), compiles it to GPU kernels, uploads it to the device, executes it, and retrieves the results.
+Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011} is a
+data-parallel array language embedded into Haskell, which targets
+GPUs. It is a \emph{generative library}, as its data-parallel array
+operations are not executed directly, but instead construct abstract
+syntax trees (AST) representing an entire data-parallel
+subcomputation. These \emph{computation representations} are executed
+using a \verb+run+ operation that accepts such an AST (of type
+\verb+Acc a+), compiles it to GPU kernels, uploads it to the device,
+executes it, and retrieves the results.\footnote{To distinguish
+  Haskell code from Agda code, we display the former in blue and with
+  a vertical bar on the left side.}
 %
-\begin{verbatim}
+\begin{hcode}
 CUDA.run :: Arrays a => Acc a -> a
-\end{verbatim}
+\end{hcode}
 %
@@
 %
-\begin{verbatim}
+\begin{hcode}
 dotp :: Vector Float -> Vector Float -> Acc (Scalar Float)
@@
              in  fold (+) 0 (zipWith (*) xs' ys')
-\end{verbatim}
+\end{hcode}
 %
@@
 %
-\begin{verbatim}
+\begin{hcode}
 type Scalar e = Array DIM0 e
 type Vector e = Array DIM1 e
-\end{verbatim}
+\end{hcode}
 %
@@
 %
-\begin{verbatim}
+\begin{hcode}
 use :: Elt e => Array sh e -> Acc (Array sh e)
-\end{verbatim}
+\end{hcode}
 %
@@
 %
-\begin{verbatim}
+\begin{hcode}
 data Z       = Z
 data sh :. i = sh :. i
-\end{verbatim}
+\end{hcode}
 %
@@
 %
-\begin{verbatim}
+\begin{hcode}
 type DIM0 = Z
 type DIM1 = DIM0 :. Int
-<and so on>
-\end{verbatim}
+-- <and so on>
+\end{hcode}
 
@@
 In this section, we investigate the potential uses of dependent typing
-in a language like Accelerate.
+in a language like Accelerate. First, we review some basics of the
+embedding.
+
+\subsection{Embedding of Haskell Types}
+\label{sec:embedding-types}
+
+Accelerate supports a wide range of numeric types as base types for
+array computations. Almost all of these types have no suitable
+counterpart in Agda, which only supports natural numbers in unary
+encoding. For that reason, our embedding keeps the Haskell types
+abstract in Agda. To specify type signatures and in particular
+functions that are polymorphic in such a Haskell type or depend on it
+in some way, we have reified these types in an Agda type
+\texttt{Element}.
+\begin{verbatim}
+data Element : Set where
+  Bool   : Element
+  Int    : Element
+  Float  : Element
+  Double : Element
+  Pair   : Element -> Element -> Element
+  -- and so on
+\end{verbatim}
+Corresponding to Haskell type classes that are used in Accelerate, our
+embedding supplies predicates that characterize subsets. For example,
+the set of numeric types is defined by a predicate \texttt{IsNumeric}.
+\begin{alltt}
+IsNumeric : Element -> Set
+IsNumeric Int = \(\top\)
+IsNumeric Float = \(\top\)
+IsNumeric Double = \(\top\)
+IsNumeric _ = \(\bot\)
+\end{alltt}
+The embedding declares further subsets all in the same style.
+
+\subsection{Array Types}
+\label{sec:array-types}
+
+To see the Agda embedding in action, we translate the
+dot product example from Section~\ref{sec:accelerate} to Agda.
+\begin{code}
+dotp : {E : Element} {{p : IsNumeric E}} {n : Nat}
+     -> PreVector n E -> PreVector n E -> Scalar E
+dotp{E} xs ys = 
+  let xs' = use xs
+      ys' = use ys
+  in
+  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
+\end{code}
+Unlike the Accelerate code, this function is polymorphic with respect
+to the array element type, provided it is numeric. It furthermore
+takes the length $n$ as a parameter thus ensuring that the two input
+vectors have the same size. The \texttt{PreVector} type of the
+arguments corresponds to the plain \texttt{Vector} type in Accelerate,
+whereas the result type \texttt{Scalar E} corresponds to \texttt{Acc
+  (Scalar E)}---a piece of abstract syntax.
+
+The \texttt{use} function works as before, but its type
+includes more information:
+\begin{code}
+use : {sh : Shape}{E : Element} -> PreArray sh E -> Array sh E
+\end{code}
+Like \texttt{E}, the index \texttt{sh} is now an element of an ordinary type:
+\begin{verbatim}
+data Shape : Set where
+  Z     : Shape
+  _:<_> : Shape -> Nat -> Shape
+\end{verbatim}
+Asking for arrays of equal shape, as in the signature of \texttt{use},
+means that the arrays have to have the exact same layout.
+The functions \texttt{fold}, \texttt{zipWith}, and \texttt{:::} are
+discussed in the subsequent Subsections. The functions \verb|_+_| and
+\verb|_*_| both have the same type:
+\begin{code}
+_+_ : {E : Element} {{p : IsNumeric E}} -> Exp E -> Exp E -> Exp E
+\end{code}
+They are restricted to arguments of numeric type and construct
+abstract syntax for an addition or a multiplication by delegating to
+the corresponding Accelerate functions.
 
@@
 the shape of the array arguments and results. These constraints ensure
-that no shape mismatches occur (e.g., a one-dimensional array is
+that no shape mismatches occur (e.g., a one-dimensional array cannot be
 considered two-dimensional). However, they do not ensure at compile
@@
 \texttt{sh'} and changes the layout of that array to \texttt{sh}. 
-\begin{verbatim}
+\begin{hcode}
 reshape :: Exp sh -> Acc (Array sh' e) -> Acc (Array sh e)
-\end{verbatim}
+\end{hcode}
 For this reshaping to work correctly, the underlying
@@
 
-In Agda, we define the shape type as follows.
-\begin{verbatim}
-data Shape : Set where
-  Z     : Shape
-  _:<_> : Shape -> Nat -> Shape
-\end{verbatim}
-The \texttt{size} function computes the number of elements stored in a
-shape.
+As \texttt{Shape} is an ordinary data type, we can define a
+\texttt{size} function that computes the number of elements stored in
+an array of a certain shape.
 \begin{verbatim}
@@
 Accelerate. 
-\begin{verbatim}
+\begin{hcode}
 fold  :: (Shape ix, Elt a) =>
@@
          Acc (Array (ix :. Int) a) -> Acc (Array ix a)
-\end{verbatim}
+\end{hcode}
 In both cases, the text of the documentation says that ``the first
@@
 
-\subsection{Embedding of Haskell Types}
-\label{sec:embedding-constants}
-
-Accelerate supports a wide range of numeric types as base types for
-array computations. Almost all of these types have no suitable
-counterpart in Agda, which only supports natural numbers in unary
-encoding. For that reason, our embedding keeps these types 
-abstract in Agda. To specify functions that are polymorphic in such a
-Haskell type or depend on it in some way, we have reified these types
-in an Agda type \texttt{Element}.
+In any case, providing such information would be done by including an
+additional argument that holds a suitable proof object, as in
 \begin{verbatim}
-data Element : Set where
-  Bool   : Element
-  Int    : Element
-  Float  : Element
-  Double : Element
-  Pair   : Element -> Element -> Element
-  ...
+fold : {E}{sh}{n} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E)
+     -> Array (sh :< n >) E -> IsMonoid f e -> Array sh E
 \end{verbatim}
-Corresponding to Haskell type classes that are used in Accelerate, our
-embedding supplies predicates that characterize subsets. For example,
-the set of numeric type is defined by a predicate \texttt{IsNumeric}.
-\begin{alltt}
-IsNumeric : Element -> Set
-IsNumeric Int = \(\top\)
-IsNumeric Float = \(\top\)
-IsNumeric Double = \(\top\)
-IsNumeric _ = \(\bot\)
-\end{alltt}
+where
+\begin{verbatim}
+IsMonoid : {E} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E) -> Set
+IsMonoid f e = ( IsAssociative f , IsUnit f e)
+\end{verbatim}
 
@@
 \begin{verbatim}
-exists (\ m : Nat -> m <= n -> Vector m E)
+exists Nat (\ m -> m <= n -> Vector m E)
 \end{verbatim}
@@
 %%% TeX-master: t
+%%% coding: utf-8
 %%% End:

### 2012-08-01 paper/main.tex
implementation

@@
 
-Agda is compiled to Haskell.
+Ordinarily, Agda is an interactive tool for constructing proofs and
+verified programs. Programs may be run, which amounts to normalizing
+Agda expressions, but this process is not very efficient.
 
-The compiler supports a Haskell foreign function interface.
+Alternatively, an interactively developed program may be compiled to
+Haskell using the Alonzo compiler. This compiler supports a Haskell
+foreign function interface (FFI), which enables Agda programs to invoke
+Haskell functions.
 
-The sophisticated type structure of the Accelerate language is in the
-way.
+Using this interface amounts to declaring a typed identifier in Agda
+and then binding the identifier to a suitably typed Haskell
+function. As an example, consider the import of the \texttt{use}
+function.
+\begin{code}
+postulate 
+  useHs : {E : Set}
+      -> HsEltDict E -> HsArray HsDIM1 E -> Acc (AccArray HsDIM1 E)
+  {-# COMPILED useHs       (\ _ -> Accel.use) #-}
+\end{code}
+The first three lines introduce the typed identifier \texttt{useHs}
+and the last line is a pragma for the Alonzo compiler that binds the
+Agda identifier \texttt{useHs} to the Haskell expression on the right.
+But wait, this type looks very unpleasant and quite different to the
+one mentioned in Section~\ref{sec:array-types}. This difference arises
+because the type translation of Alonzo is unable to cope with the
+index type \texttt{Shape}. For that reason, the interface uses a
+simplified array type and adapter functions are required, both on the
+Agda side and on the Haskell side of the interface.
 
-The foreign function interface is not sufficiently expressive so that
-Agda types need to be encoded and corresponding structures need to be
-build from this encoding in the Haskell code.
+At the foreign function interface level, all arrays are considered as
+one-dimensional arrays. Additional arguments are passed to encode the
+additional structure. The Agda adapter provides the encoding of this
+structure and the Haskell adapter decodes it again.
+
+We believe that these adapations only have a minor performance impact
+because (1) most functions just manipulate abstract syntax, so that
+only AST construction is affected, and (2) internally, Accelerate
+considers all arrays as one-dimensional so that operations like
+\texttt{reshape} are no-ops at run time.
+
+Here is the Agda adapter for \texttt{use}:
+\begin{verbatim}
+use : {sh : Shape}{E : Element} -> PreArray E sh -> Array E sh
+use {sh}{E} (PA y) = Ar (useHs (EltDict E) y)
+\end{verbatim}
+It makes use of two wrapper types. \texttt{PreArray} wraps a
+one-dimensional Haskell using the constant \texttt{HsDIM1} (the
+\texttt{DIM1} type shown in Section~\ref{sec:accelerate} imported from
+Haskell via FFI) and the function \texttt{EltType} (not shown), which
+interprets a value of type \texttt{Element} as a Haskell type. The
+latter types are also imported via FFI.
+\begin{code}
+data PreArray (Elt : Element) : Shape -> Set where
+  PA : {sh : Shape}
+     -> HsArray HsDIM1 (EltType Elt) -> PreArray Elt sh
+\end{code}
+The \texttt{Array} type wraps an AST reference for an Accelerate
+array, where \texttt{Acc} and \texttt{AccArray} are types imported
+from Haskell.
+\begin{code}
+data Array (Elt : Element) : Shape -> Set where
+  Ar : {sh : Shape}
+     -> Acc (AccArray HsDIM1 (EltType Elt)) -> Array Elt sh
+\end{code}
+The \texttt{EltDict} function translates a value 
+\texttt{(E : Element)} into a Haskell expression that evaluates to a
+dictionary for the Haskell type of \texttt{E} for the Haskell type
+class \texttt{Elt}. Such a dictionary is passed, whenever that
+corresponding Haskell function has type class constraints.
+\begin{code}
+EltDict : (E : Element) -> HsEltDict (EltType E)
+\end{code}
+
+The Haskell side of the adapter has several purposes. First, it
+materializes the type class dictionaries from the encoding that we
+just discussed. Second, it reconstructs sufficient information about
+the array shape so that the intended operation can execute. Here is
+the code for \texttt{Accel.use}, where the module name \texttt{A}
+is a shorthand for \texttt{Data.Array.Accelerate}.
+\begin{hcode}
+use :: EltDict e -> Array A.DIM1 e -> A.Acc (A.Array A.DIM1 e)
+use EltDict (ARRAY ar) = (A.use ar)
+\end{hcode}
+It does not have to reconstruct any information except the type class
+constraint. This constraint is materialized using the type
+\texttt{EltDict} below.
+\begin{hcode}
+data EltDict e where
+  EltDict :: (A.Elt e) => EltDict e
+\end{hcode}
+This datatype is built such that each value captures the \texttt{Elt}
+dictionary of type \texttt{e}. It remains to build such values for all
+types that we want to transport across the FFI. These are the values
+used by the (Agda) \texttt{EltDict} function. Here are two examples.
+\begin{hcode}
+eltDictBool :: EltDict Bool
+eltDictBool = EltDict
+
+eltDictInt :: EltDict Int
+eltDictInt = EltDict
+\end{hcode}
+
+As an example for a function that require more work on either side,
+consider the \texttt{fold} operation. 
+\begin{code}
+fold : {E}{sh}{n}
+     -> (Exp E -> Exp E -> Exp E)
+     -> Exp E
+     -> Array E (sh :< n >)
+     -> Array E sh
+fold {E}{sh}{n} f (Ex e) (Ar a) =
+  Ar (foldHs (EltDict E) (toHsInt (size sh)) (toHsInt n)
+             (unwrap2 f) e a)
+\end{code}
+As values of type \texttt{Exp} also need a wrapper type in Agda (it is
+not possible to import type constructors via the FFI), there is some
+unwrapping going on for the \texttt{e} and \texttt{f}
+arguments. The implementation of \texttt{fold} just calls the
+\texttt{foldHs} function and encodes the information about the shape
+in two number arguments. Here, \texttt{size sh} is the size of the
+result and \texttt{n} is the size of the dimension that is folded. As
+these values are first available as Agda natural numbers, they need to
+be converted to Haskell numbers using the function \texttt{toHsInt}.
+
+The \texttt{foldHs} function is defined via the FFI.
+\begin{code}
+postulate
+  foldHs : {A : Set}
+          -> HsEltDict A
+          -> HsInt
+          -> HsInt
+          -> (AccExp A -> AccExp A -> AccExp A)
+          -> AccExp A
+          -> Acc (AccArray HsDIM1 A)
+          -> Acc (AccArray HsDIM1 A)
+  {-# COMPILED foldHs      (\_ -> Accel.fold) #-}
+\end{code}
+The Haskell adapter needs to reconstruct the \texttt{Elt} dictionary
+as before, but it also needs to reshape the one-dimensional array
+representation into a two-dimensional one for executing the fold
+operation. The two size arguments are required for exactly this
+reshape operation. With that insight, the code is straightforward.
+\begin{hcode}
+fold :: EltDict a
+     -> Int -> Int
+     -> (A.Exp a -> A.Exp a -> A.Exp a)
+     -> A.Exp a
+     -> A.Acc (A.Array A.DIM1 a)
+     -> A.Acc (A.Array A.DIM1 a)
+fold EltDict size2 size1 f e a =
+     (A.reshape (A.lift (A.Z A.:. size2))
+      (A.fold f e
+       (A.reshape (A.lift (A.Z A.:. size2 A.:. size1)) a)))
+\end{hcode}
+
+Fortunately, the \texttt{fold} example is about as complicated as the
+adapter code gets. There are also many cases where at least one side
+of the adapter code is trivial. However, each case must be considered
+separately.
 

### 2012-08-01 paper/main.tex
conclusion

@@
 
+We have build an experimental Agda frontend for the Accelerate
+language. The goal of this experiment was to explore potential uses of
+dependently-typed programming for data-parallel languages.
+
+At the moment, the outcome of the experiment is mixed. It is
+successful, because we have been able to construct Agda functions for
+a representative sample of Accelerate's functionality.
+However, there was less scope for encoding extra information in the
+dependent types than we had hoped for. Exact matching of array bounds
+works, but results in restrictions (like the problems with
+\texttt{zipWith} and filtering) that were not anticipated.
+
+Exploiting algebraic properties did not work out in the intended way, mainly
+because it boils down to asserting that some AST denotes an
+associative function. However, these assertions cannot be proven: the
+proof would have to apply the semantics to the AST, but the AST is an
+abstract type in our implementation. An AST representation in Agda
+might give us a better handle at this problem.
+
+In some places, the Agda frontend is less dynamic than Accelerate. In
+a number of places, Accelerate accepts a run-time value of type
+\texttt{Exp sh} for a shape argument, where the Agda frontend requires
+a value of type \texttt{Shape}. To address this problem, we would have
+to include a \texttt{Shape}-indexed encoding of the \texttt{Shape}
+type in the \texttt{Element} type so that we can describe the type of
+an expression whose value has a certain shape.
+
+Finally, the type translation of Agda's FFI has many shortcomings that
+caused a number of problems for transporting information between Agda
+and Haskell. One part of the problem is, unfortunately, the rich type
+structure of Accelerate which already encodes much useful
+information. An alternative, untyped (or less-typed) interface to
+Accelerate would make the adaptation to an Agda frontend much simpler.
+
 %

### 2012-08-13 paper/main.tex
Small changes throughout

@@
 Instead, GPUs require a high degree of data parallelism, restricted control flow, and carefully tailored data access patterns to be efficient.
-Previous work ---for example, Accelerator~\cite{??}, Copperhead~\cite{??}, and Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011}--- demonstrates that embedded array languages with a custom code generator can meet those GPU constraints by carefully limiting the embedded language constructs.
+Previous work ---for example, Accelerator~\cite{Tarditi:2006}, Copperhead~\cite{Catanzaro:EECS-2010-124}, and Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011}--- demonstrates that embedded array languages with a custom code generator can meet those GPU constraints by carefully limiting the embedded language constructs.
 
@@
 
-This paper is a first investigation into this topic. It reports of a
+This paper is a first investigation into this topic. It reports on a
 partial port of Accelerate to a new, dependently-typed host language,
@@
 This code cannot fail at run time because a caller has to
-construct the proof tree for \verb+m < n+ before invoking \verb+get+.
+construct the proof tree for \verb+m < n+ before invoking \verb+get+. (In Agda, arguments in curly braces are \emph{implicit arguments} that will be inferred if omitted in an application.)
 
@@
 subcomputation. These \emph{computation representations} are executed
-using a \verb+run+ operation that accepts such an AST (of type
+using a \verb+run+ operation that accepts such an representation (of type
 \verb+Acc a+), compiles it to GPU kernels, uploads it to the device,
@@
 Unlike the Accelerate code, this function is polymorphic with respect
-to the array element type, provided it is numeric. It furthermore
+to the array element type, provided it is numeric.\footnote{In Agda, arguments in double curly braces are \emph{instance arguments}~\cite{DevriesePiessens2011}.
+We use them much like type class constraints are used in Haskell.} It furthermore
 takes the length $n$ as a parameter thus ensuring that the two input
@@
 The functions \texttt{fold}, \texttt{zipWith}, and \texttt{:::} are
-discussed in the subsequent Subsections. The functions \verb|_+_| and
+discussed in the subsequent subsections. The functions \verb|_+_| and
 \verb|_*_| both have the same type:
@@
 to reshape a two-dimensional $3\times 4$-array to a vector of size
-$12$ or to a three-dimensional $3\times2\times2$-array, whereas an attempt to
+$12$ or to a three-dimensional $3\times2\times2$-array, an attempt to
 reshape the same array to a $2\times5$-array should be rejected at
@@
 remaining ones are to be inferred by Agda.
-The argument \texttt{nu} is an instance argument that is automatically
-filled-in with a suitably typed value in scope
+As mentioned before, the argument \texttt{nu} is an instance argument; it is automatically
+filled-in with a suitably typed value that is currently in scope
 \cite{DevriesePiessens2011}. Here, the predicate 
@@
 applicability of dependent typing. We already mentioned that the
-formalization of associativity or of the concept monoid gets
+formalization of associativity or of the concept of a monoid gets
 unmanageable because such properties have to be asserted for abstract syntax.
@@
 \texttt{m} is that it must be less than or equal to
-\texttt{n}. However, we cannot prove this from the code because the
-Accelerate implementation computes the length of the result in a quite
-subtle way. Moreover, as Accelerate deals with the construction of
-abstract syntax most of the time, the value of \texttt{m} is simply
-not available at run time of \texttt{filter}.
+\texttt{n}.
+However, we cannot prove this from the code because of a staging restriction.
+The Accelerate implementation computes the length of the result only when the generated GPU code is executed.
+The function \texttt{Exp E -> Exp Bool} maps abstract syntax to abstract syntax; it does not directly implement a Boolean predicate.
+Hence, we cannot use the predicate in the result type of \texttt{filter} to more accurately constrain the size of the resulting vector --- unless we include an evaluator for Accelerate expressions.
 
-One might contemplate that an existential type like
+We might contemplate emplying an existential type like
 \begin{verbatim}
@@
 \end{verbatim}
-could be employed. However, it is not possible to build such an
+However, it is not possible to build such an
 existential package because the evidence \texttt{m} is not available
@@
 There are several ways of implementing this idea. The simplest
-approach is to wrap each element in a maybe type.
+approach is to wrap each element in a maybe type or
 pair up each element with a boolean flag that indicates
-its presence.\footnote{A \texttt{Maybe} type would be a better option,
+its presence.\footnote{A \texttt{Maybe} type is on one hand the better option,
   but it has to be coded without using pattern matching.}

### 2012-08-15 paper/main.tex
final changes for the draft proceedings

@@
 
-Generative approaches to programming parallel hardware promise to combine high-level programming models with high-performance.
-They are especially attractive for targeting restricted architectures, such as GPUs (graphics processor units), that cannot efficiently execute code aimed at conventional multicore CPUs.
-Instead, GPUs require a high degree of data parallelism, restricted control flow, and carefully tailored data access patterns to be efficient.
-Previous work ---for example, Accelerator~\cite{Tarditi:2006}, Copperhead~\cite{Catanzaro:EECS-2010-124}, and Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011}--- demonstrates that embedded array languages with a custom code generator can meet those GPU constraints by carefully limiting the embedded language constructs.
+Generative approaches to programming parallel hardware promise to
+combine high-level programming models with high performance. They are
+particularly attractive for targeting restricted architectures, such
+as GPUs (graphics processor units), that cannot efficiently execute
+code aimed at conventional multicore CPUs. Instead, GPUs require a
+high degree of data parallelism, restricted control flow, and
+carefully tailored data access patterns to be efficient. Previous work
+---for example, Accelerator~\cite{Tarditi:2006},
+Copperhead~\cite{Catanzaro:EECS-2010-124}, and
+Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011}---
+demonstrates that embedded array languages with a custom code
+generator can meet those GPU constraints with carefully designed language constructs.
 
-Given a host languages with an expressive type system, it is attractive to leverage that type system to express static properties of the embedded language.
-For example, Accelerate, an embedded array language for Haskell, uses Haskell's recent support for type-level programming like GADTs and type families in that manner~\cite{ChakravartyKellerLeeMcdonellGrover2011}.
-This is especially important for approaches relying on runtime code generation as we want to mimise the number of potential faults at application runtime, which coincides with the compile time of the embedded language.
-Moreover, static guaranties hold the potential to improve the predictability of parallel performance.
+Given a host languages with an expressive type system, it is
+attractive to leverage that type system to express static properties
+of the embedded language. For example, Accelerate, an embedded array
+language for Haskell, uses Haskell's recent support for type-level
+programming like GADTs and type families in that
+manner~\cite{ChakravartyKellerLeeMcdonellGrover2011}. This design
+choice is important for approaches relying on runtime code generation:
+compile-time faults in the embedded language should be avoided because
+it corresponds to at application runtime. Moreover, static guaranties
+hold the potential to improve the predictability of parallel
+performance.
 
-Dependent types are an emerging approach to certified programming, where invariants are established in the form of types and proved at compile time.
+Dependent types are an emerging approach to certified programming, where invariants are established in the form of types and proven at compile time.
 Many of Haskell's type-level extensions used in Accelerate approximate various aspects of dependently-typed programming.
@@
 suited to this port because of its foreign function interface to
-Haskell, which enables it to directly invoke functionality of
+Haskell, which enables it to directly invoke the functionality of
 Accelerate.
@@
 dependent types in an array-oriented data parallel language like
-Accelerate. Section~\ref{sec:limitations} considers conceptual
+Accelerate and how they were realized in our
+implementation. Section~\ref{sec:limitations} considers conceptual 
 problems and limitations that we ran into when constructing the Agda
@@
 
-A familiar example for an indexed data type is the type
-\verb+Vec A n+ for vectors of fixed length \verb+n+ and elements of
-type \verb+A+ program program with an access operation that restricts the index to the
+One attractive feature of Agda's inductive data type families is the
+ability to construct indexed data types.
+A familiar example for such an indexed data type is the type
+\verb+Vec A n+ of vectors of fixed length \verb+n+ and elements of
+type \verb+A+. This vector data types can be equipped with an access
+operation that restricts the index to the 
 actual length of the vector at compile time.\footnote{An
@@
 \end{verbatim}
-The above defines an indexed data type \verb+Vec A n+ with two
+The above defines the type \verb+Nat+ of natural numbers and an
+indexed data type \verb+Vec A n+ where \texttt{A} is a type and
+\texttt{n} is a natural number. The latter type comes with two 
 constructors, \verb+[]+ for the vector of length zero and 
@@
 This code cannot fail at run time because a caller has to
-construct the proof tree for \verb+m < n+ before invoking \verb+get+. (In Agda, arguments in curly braces are \emph{implicit arguments} that will be inferred if omitted in an application.)
+construct the proof tree for \verb+m < n+ before invoking
+\verb+get+. (In Agda, arguments in curly braces are \emph{implicit
+  arguments} that will be inferred if omitted in an application.)
+Thus, an ``index out of bounds'' error cannot happen.
 
@@
 subcomputation. These \emph{computation representations} are executed
-using a \verb+run+ operation that accepts such an representation (of type
+using a \verb+run+ operation that accepts such a representation (of type
 \verb+Acc a+), compiles it to GPU kernels, uploads it to the device,
 executes it, and retrieves the results.\footnote{To distinguish
-  Haskell code from Agda code, we display the former in blue and with
+  Haskell code from Agda code, we display Haskell code in blue and with
   a vertical bar on the left side.}
@@
 
-The type \verb+Scalar+ and \verb+Vector+ are merely type synonyms instantiating a shape-parameterised array type to the special case of zero and one dimensional arrays:
+The types \verb+Scalar+ and \verb+Vector+ are type synonyms instantiating a shape-parameterised array type to the special case of zero and one dimensional arrays:
 %
@@
 \end{hcode}
-%
-So, the general type for \verb+use+ is
+
+The general type for \verb+use+ is
 %
@@
 
-Common dimensions, such as \verb+DIM0+, \verb+DIM1+, and so on, are predefined, but to enable shape polymorphic computations, along the lines pioneered in the Haskell array library Repa~\cite{keller-etal:repa}, shapes are inductively defined using type-level snoc lists formed from
+Common dimensions, such as \verb+DIM0+, \verb+DIM1+, and so on, are
+predefined, but to enable shape polymorphic computations, along the
+lines pioneered in the Haskell array library
+Repa~\cite{keller-etal:repa}, shapes are inductively defined using
+type-level snoc lists built from the data types \texttt{Z} and \texttt{:.}:
 %
@@
 %
-Hence, we have got
+Hence, the definitions of the dimensions:
 %
@@
 
-Dependent ML is an ML dialect with restricted form of dependent types, which, among other applications, may be used to statically check array bounds in array operations~\cite{xi:dml-jfp}.
+Dependent ML is an ML dialect with a restricted form of dependent types, which, among other applications, may be used to statically check array bounds in array operations~\cite{xi:dml-jfp}.
 However, only simple indexing and array updating are considered and not aggregate array operations, such as those provided by Accelerate.
@@
 In this section, we investigate the potential uses of dependent typing
-in a language like Accelerate. First, we review some basics of the
+in a language like Accelerate and point out how they may be
+implemented in Agda. First, we review some basics of the
 embedding.
@@
 
-Accelerate supports a wide range of numeric types as base types for
+Accelerate supports a wide range of numeric types, characterized by
+type class \texttt{Elt}, as base types for
 array computations. Almost all of these types have no suitable
@@
 embedding supplies predicates that characterize subsets. For example,
-the set of numeric types is defined by a predicate \texttt{IsNumeric}.
+the set of numeric types is defined by a predicate
+\texttt{IsNumeric}.\footnote{$\top$ is a one-element type,
+  whereas $\bot$ is a type without elements. These types customarily
+  represent truth and falsity.}
 \begin{alltt}
@@
 \end{code}
-Like \texttt{E}, the index \texttt{sh} is now an element of an ordinary type:
+Like \texttt{E}, the index \texttt{sh} is now an element of an
+ordinary type instead of having to rely on type-level snoc
+lists:\footnote{Recent work on Haskell's type system manages to avoid
+  this issue \cite{YorgeyWeirichCretinJonesVytiniotisMagalhaes2012}.}
 \begin{verbatim}
@@
 Asking for arrays of equal shape, as in the signature of \texttt{use},
-means that the arrays have to have the exact same layout.
+means that the arrays have to have the exact same layout. The
+\texttt{PreVector} and \texttt{Vector} types are just synonyms as in
+Haskell:
+\begin{verbatim}
+PreVector n E = PreArray (Z :< n >) E
+Vector n E    = Array    (Z :< n >) E
+\end{verbatim}
 The functions \texttt{fold}, \texttt{zipWith}, and \texttt{:::} are
@@
 abstract syntax for an addition or a multiplication by delegating to
-the corresponding Accelerate functions.
+the corresponding Accelerate functions. The type \texttt{Exp E},
+which denotes an AST of an expression of type \texttt{E}.
 
@@
 $12$ or to a three-dimensional $3\times2\times2$-array, an attempt to
-reshape the same array to a $2\times5$-array should be rejected at
+reshape to a $2\times5$-array should be rejected at
 compile time.
 
-As \texttt{Shape} is an ordinary data type, we can define a
+As \texttt{Shape} is an ordinary data type in Agda, we can define a
 \texttt{size} function that computes the number of elements stored in
@@
 involves an extra argument with a proof that the source and
-target shapes have the same size.\footnote{For the moment, the type
-  \texttt{Element} could be read as \texttt{Set}. Actually,
-  \texttt{Element} classifies the types of values that can be stored
-  in Accelerate arrays. In Haskell, this restriction is imposed with a
-  type class.}
-\begin{verbatim}
-reshape : {sh : Shape} {E : Element}
-       -> (sh' : Shape) -> Array sh E -> (size sh = size sh')
+target shapes have the same size.
+\begin{alltt}
+reshape : \{sh : Shape\} \{E : Element\}
+       -> (sh' : Shape) -> Array sh E -> (size sh \(\equiv\) size sh')
        -> Array sh' E
-\end{verbatim}
+\end{alltt}
 There is a subtle difference to the original signature. In
@@
 The latter type is more restrictive than the Accelerate
-implementation of \texttt{zipWith}. Instead of checking the dimensions
+implementation of \texttt{zipWith}. Instead of checking the sizes
 of the input arrays, it truncates them to the respective minima. A
-corresponding Agda type could be developed easily. It would involve
-specifying a suitable ternary relation between shapes.
+corresponding Agda type could be developed easily. It just requires a
+binary function that computes the minimum of two shapes, which is a
+simple exercise.
 
@@
 \begin{verbatim}
-fold1 : ... -> Array (sh :< n >) E -> (p : size sh * n > 0)
+fold1 : ... -> Array (sh :< n >) E -> (size sh * n > 0)
             -> Array sh E
@@
 Accelerate relies on Haskell's built-in support for the type classes
-\texttt{Num} and \texttt{Fractional} to embed constants. The Haskell compiler reads each integer
-literal as a value of type \texttt{Integer}, which is a built-in type
-of arbitrary precision integers. To this value, Haskell applies the
-function \texttt{fromInteger} that converts to the expected
-type. Similarly, floating point constants are read as values of
+\texttt{Num} and \texttt{Fractional} to embed constants. The Haskell
+compiler reads each integer literal as a value of type
+\texttt{Integer}, which is a built-in type of arbitrary precision
+integers. To this value, Haskell applies the function
+\texttt{fromInteger} that converts to the type expected by the
+context. Similarly, floating point constants are read as values of
 type \texttt{Rational} (\texttt{Integer} fractions) and then converted
-using \texttt{fromRational}. Accelerate provides instances of
-these type classes that define \texttt{fromInteger} and
+using \texttt{fromRational}. Accelerate provides instances of these
+type classes that define \texttt{fromInteger} and
 \texttt{fromRational} to produce suitable AST fragments.
@@
 
-For a similar problem, consider an implementation of the \texttt{filter}
+For a related problem, consider an implementation of the \texttt{filter}
 operation that takes a predicate and a source array and returns an
@@
 existential package because the evidence \texttt{m} is not available
-when the package is constructed.
+when the package would have to be constructed.
 
@@
 
-Now filtering becomes quite simple.
+Now filtering becomes quite simple because the length of
+the\texttt{FVector} does not change. Furthermore, filtering could be
+extended to multi-dimensional arrays, although it is not clear if an
+array that includes non-existing elements is a sensible notion.
 \begin{verbatim}
@@
 index type \texttt{Shape}. For that reason, the interface uses a
-simplified array type and adapter functions are required, both on the
-Agda side and on the Haskell side of the interface.
+simplified array type and adapter functions are required, in the worst
+case, both on the Agda side and on the Haskell side of the interface.
 
@@
 one-dimensional arrays. Additional arguments are passed to encode the
-additional structure. The Agda adapter provides the encoding of this
-structure and the Haskell adapter decodes it again.
+shape information as far as it is needed. The Agda adapter provides
+the encoding of this structure and the Haskell adapter decodes it again.
 
@@
 \begin{verbatim}
-use : {sh : Shape}{E : Element} -> PreArray E sh -> Array E sh
+use : {sh : Shape}{E : Element} -> PreArray sh E -> Array sh E
 use {sh}{E} (PA y) = Ar (useHs (EltDict E) y)
@@
 It makes use of two wrapper types. \texttt{PreArray} wraps a
-one-dimensional Haskell using the constant \texttt{HsDIM1} (the
+one-dimensional Haskell array using the constant \texttt{HsDIM1} (the
 \texttt{DIM1} type shown in Section~\ref{sec:accelerate} imported from
@@
 \begin{code}
-data PreArray (Elt : Element) : Shape -> Set where
-  PA : {sh : Shape}
-     -> HsArray HsDIM1 (EltType Elt) -> PreArray Elt sh
+data PreArray (E : Element) : Shape -> Set where
+  PA : {sh : Shape} -> HsArray HsDIM1 (EltType E) -> PreArray sh E
 \end{code}
@@
 \begin{code}
-data Array (Elt : Element) : Shape -> Set where
-  Ar : {sh : Shape}
-     -> Acc (AccArray HsDIM1 (EltType Elt)) -> Array Elt sh
+data Array (E : Element) : Shape -> Set where
+  Ar : {sh : Shape} -> Acc (AccArray HsDIM1 (EltType E)) -> Array sh E
 \end{code}
@@
 
-As an example for a function that require more work on either side,
+As an example for a function that requires more work on either side,
 consider the \texttt{fold} operation. 
@@
      -> Exp E
-     -> Array E (sh :< n >)
-     -> Array E sh
+     -> Array (sh :< n >) E
+     -> Array sh E
 fold {E}{sh}{n} f (Ex e) (Ar a) =
@@
 \texttt{foldHs} function and encodes the information about the shape
-in two number arguments. Here, \texttt{size sh} is the size of the
+in two integer arguments. Here, \texttt{size sh} is the size of the
 result and \texttt{n} is the size of the dimension that is folded. As
-these values are first available as Agda natural numbers, they need to
+these values are initially available as Agda natural numbers, they need to
 be converted to Haskell numbers using the function \texttt{toHsInt}.
@@
 \end{code}
-The Haskell adapter needs to reconstruct the \texttt{Elt} dictionary
+The Haskell adapter reconstructs the \texttt{Elt} dictionary
 as before, but it also needs to reshape the one-dimensional array

### 2012-08-30 paper/main.tex
draft of talk

@@
 
-Given a host languages with an expressive type system, it is
+Given a host language with an expressive type system, it is
 attractive to leverage that type system to express static properties

### 2012-08-30 talk/ifl2012.tex
draft of talk

\documentclass{beamer}
% 20+5 minutes
\input{macros}

\AtBeginSection[]
{
  \begin{frame}<beamer>
    \frametitle{Outline}
    \tableofcontents[currentsection,currentsubsection]
  \end{frame}
}

\title{Agda Meets Accelerate}
\date{1.9.2012}
\author{Peter Thiemann \and Manuel Chakravarty}
\institute{%
  University of Freiburg, Germany,\\
  \texttt{thiemann@informatik.uni-freiburg.de}
\and
University of New South Wales, Sydney, Australia,\\
\texttt{chak@cse.unsw.edu.au}
}
\begin{document}

% begin titlepage
% "`normal"' front page template:
%\setbeamertemplate{title page}[default][rounded=true]
\setbeamertemplate{navigation symbols}{}
\begin{frame}[plain,label=fp]
		\advance\textwidth by 3.2em\relax
		\begin{minipage}{\textwidth}\par%
				\maketitle
		\end{minipage}
		\hspace*{2.5em}%
\end{frame}
\makeatother 
% end titlepage
\setbeamertemplate{navigation symbols}[horizontal]

\section{Introduction}

\begin{frame}
  \frametitle{Introduction}
  \begin{itemize}
  \item Embedded languages
  \item Dependent types
  \item Generative programming
  \item Data-parallel programming
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Overview}
  \begin{block}{Mission}
    Construct a generative, data parallel programming language embedded in Agda
  \end{block}
  \begin{block}<2->{Means}
    \dots\ by creating a --- more expressive --- Agda frontend for the
    Accelerate language.
  \end{block}
\end{frame}

\section{Accelerate}
\begin{frame}
  \frametitle{The Accelerate Language}
  \begin{itemize}
  \item Data-parallel array language
  \item Expressive embedding in Haskell
  \item Generates GPU code at run time
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Accelerate Example}
  \framesubtitle{Dot Product}
\begin{verbatim}
dotp :: Vector Float -> Vector Float -> Acc (Scalar Float)
dotp xs ys = let xs' = use xs
                 ys' = use ys
             in  fold (+) 0 (zipWith (*) xs' ys')
\end{verbatim}
  \begin{itemize}
  \item \texttt{dotp xs ys} returns a representation of code that
    computes the dot product of \texttt{xs} and \texttt{ys}
  \item References to inputs via \texttt{use}
  \item Specification of operations via HOAS (viz.\ \texttt{(+)} and \texttt{(*)})
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Running Accelerate Code}
  \begin{itemize}
  \item To execute \texttt{dotp}, Accelerate provides several
    implementations of a \texttt{run} function:
\begin{verbatim}
CUDA.run :: Arrays a => Acc a -> a
\end{verbatim}
  \item This implementation
    \begin{itemize}
    \item compiles the code to GPU kernels
    \item loads the \texttt{use}d arguments along with the code onto
      the GPU
    \item executes and
    \item retrieves the result.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Types in Accelerate}
  \begin{itemize}
  \item \texttt{Vector} and \texttt{Scalar} are suitable
    instantiations of a parameterized array type:
\begin{verbatim}
type Scalar e = Array DIM0 e
type Vector e = Array DIM1 e
\end{verbatim}
  \item Dimensions in turn are abbreviations for certain array shapes:
\begin{verbatim}
type DIM0 = Z
type DIM1 = DIM0 :. Int
-- <and so on>
\end{verbatim}
  \item and simple shapes are built from two basic constructors
\begin{verbatim}
data Z       = Z
data sh :. i = sh :. i
\end{verbatim}
  \end{itemize}
\end{frame}

\section{Agda}
\begin{frame}
  \frametitle{Agda}
  \begin{itemize}
  \item A dependently-typed functional programming language
    \begin{itemize}
    \item Inductive datatype families with dependent pattern matching
    \item Dependent records and parameterized modules
    \item Universe polymorphism
    \item \dots
    \end{itemize}
  \item Interactive proof assistant based on intuitionistic type theory
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Indexed Types}
  \begin{itemize}
  \item Standard example: vector type indexed by its size
\begin{verbatim}
data Vec (A : Set) : Nat -> Set where
  []   : Vec A zero
  _::_ : {n : Nat} -> A -> Vec A n -> Vec A (suc n)
\end{verbatim}
  \item Enables the construction of statically safe access operations
    for vectors
\begin{verbatim}
get : forall {A : Set} {n : Nat} ->
      Vec A n -> (m : Nat) -> m < n -> A
get []        _       ()      -- impossible case
get (x :: xs) zero    p       = x
get (x :: xs) (suc m) (s<s p) = get xs m p
\end{verbatim}
  \end{itemize}
\end{frame}

\section{Dependent Types for Accelerate}

\begin{frame}
  \frametitle{Dependent Types for Accelerate}
  \begin{block}{Expectations}
    \begin{enumerate}
    \item Augment array shapes with static size information
    \item Correctness guarantees for fold-like operations
    \item Express all Haskell examples with stronger guarantees
    \end{enumerate}
  \end{block}
  \begin{block}<2->{Results}
    \begin{enumerate}
    \item Bounds checking works out.
    \item Not satisfactory.
    \item Some workarounds required.
    \end{enumerate}
  \end{block}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Example: Dot Product in Agda-Accelerate}
  \vspace{-\baselineskip}
\begin{verbatim}
dotp : forall {E : Element} {{p : IsNumeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in
  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{verbatim}
  \begin{itemize}
  \item \texttt{PreVector n E} : a vector of $n$ elements of
    type \texttt{E}
  \item \texttt{Scalar E} : code that evaluates to a scalar of type \texttt{E}
  \item \texttt{IsNumeric E} restricts \texttt{E} to numeric types
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{The Array Type}
  \begin{itemize}
  \item The array type is indexed over a shape
  \item Shapes contain size information
\begin{verbatim}
data Shape : Set where
  Z     : Shape
  _:<_> : Shape -> Nat -> Shape
\end{verbatim}
  \item \texttt{Vector} and \texttt{Scalar} are, again,
    specializations of a shape-indexed array type
\begin{verbatim}
Vector n E    = Array    (Z :< n >) E
Scalar E      = Array    Z          E
\end{verbatim}
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{Bounds Checking: zipWith}
\begin{verbatim}
zipWith : forall {A B C} {sh} -> (Exp A -> Exp B -> Exp C)
        -> Array sh A -> Array sh B -> Array sh C
\end{verbatim}
  \begin{itemize}
  \item Ensures that only arrays of the exact same shape and size are
    zipped together
  \item More restrictive than Accelerate that only checks the
    dimension and then takes the minimum
  \item But that can also be encoded
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{Accelerate's zipWith Function in Agda}
\begin{verbatim}
zipWith' : forall {A B C} {sha shb} {p : EqShape sha shb}
        -> (Exp A -> Exp B -> Exp C)
        -> Array sha A -> Array shb B
        -> Array (shmin sha shb {p}) C
\end{verbatim}
where\scriptsize
\begin{verbatim}
data EqShape : Shape -> Shape -> Set where
  EqZ : EqShape Z Z
  EqS : {s1 s2 n1 n2} -> EqShape s1 s2 -> EqShape (s1 :< n1 >) (s2 :< n2 >)

shmin : (s1 : Shape) -> (s2 : Shape) -> { p : EqShape s1 s2 } -> Shape
shmin .Z .Z {EqZ} = Z
shmin .(s1 :< n1 >) .(s2 :< n2 >) {EqS {s1} {s2} {n1} {n2} y} =
    shmin s1 s2 {y} :< min n1 n2 >
\end{verbatim}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Bounds Checking: fold and fold1}

  \begin{itemize}
  \item \texttt{fold} and \texttt{fold1} take a binary operation and
    reduce one dimension of an array using the operation
  \item \texttt{fold} works for any array, but \texttt{fold1} requires
    a non-empty array argument
  \item The latter restriction is not expressed in Haskell-Accelerate
  \end{itemize}
  \begin{block}<2->{\vspace{-\baselineskip}}\small
\begin{verbatim}
fold  :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) -> Exp a ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
fold1 :: (Shape ix, Elt a) =>
         (Exp a -> Exp a -> Exp a) ->
         Acc (Array (ix :. Int) a) -> Acc (Array ix a)
\end{verbatim}
  \end{block}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Agda Signatures for fold and fold1}
\begin{verbatim}
fold : forall {E}{sh}{n}
     -> (Exp E -> Exp E -> Exp E) -> Exp E
     -> Array (sh :< n >) E 
     -> Array sh E

fold1 : forall {E}{sh}{n}
      -> (Exp E -> Exp E -> Exp E)
      -> Array (sh :< suc n >) E
      -> Array sh E
\end{verbatim}
\end{frame}
\begin{frame}[fragile]
  \frametitle{Another Issue with fold and fold1}
  \begin{itemize}
  \item \texttt{fold f e} and \texttt{fold1 f}  partition their array
    argument and apply \texttt{f} in a bracketing that is a-priori unknown
  \item For deterministic results, \texttt{f} must be associative and
    (in case of \texttt{fold}) \texttt{e} must be a unit of \texttt{f}
  \item<2-> Such restrictions can be expressed in Agda
  \item<3-> Might ask for a signature like this
\begin{verbatim}
fold : ... -> Monoid E
     -> Array (sh :< n >) E -> Array sh E
\end{verbatim}
  \end{itemize}
\end{frame}
\begin{frame}[fragile]
  \frametitle{A Monoidal Catch}
An encoding of monoids in Agda:\scriptsize
\begin{verbatim}
record Monoid (E : Set) : Set where
  field
    op : E -> E -> E
    un : E
    op_assoc : forall {x y z} -> op (op x y) z = op x (op y z)
    op_un_le : forall {x} -> op un x = x
    op_un_re : forall {x} -> op x un = x
\end{verbatim}
\normalsize
\begin{block}<2->{The Catch}
  In the signature of \texttt{fold} \dots
  \begin{itemize}
  \item \texttt{E} is \textbf{not} a \texttt{Set}, but an element of a
    data type that encodes Haskell datatypes
  \item the type of the operation is
    \texttt{(Exp E -> Exp E -> Exp E)}\\
    a HOAS encoding of the operation's \textbf{syntax}
  \item<3-> Associativity of syntax is meaningless \dots 
  \end{itemize}

\end{block}
\end{frame}
\end{document}

### 2012-08-30 talk/ifl2012.tex
draft

@@
   \begin{itemize}
-  \item Embedded languages
+  \item Domain-specific embedded languages
   \item Dependent types
-  \item Generative programming
-  \item Data-parallel programming
+  \item D-Generative programming
+  \item Data parallelism
   \end{itemize}
@@
   \begin{block}{Mission}
-    Construct a generative, data parallel programming language embedded in Agda
+    Construct a dependently-typed, generative, data parallel programming language
   \end{block}
   \begin{block}<2->{Means}
-    \dots\ by creating a --- more expressive --- Agda frontend for the
-    Accelerate language.
+    \dots\ by creating an Agda frontend for the Accelerate language.
   \end{block}
@@
   \item Generates GPU code at run time
+  \item See Chakravarty et al, DAMP2011
   \end{itemize}
@@
 \begin{verbatim}
-dotp :: Vector Float -> Vector Float -> Acc (Scalar Float)
+dotp :: Vector Float -> Vector Float
+     -> Acc (Scalar Float)
 dotp xs ys = let xs' = use xs
@@
 \end{verbatim}
-  \item and simple shapes are built from two basic constructors
+  \item Shapes are built from two basic constructors
 \begin{verbatim}
@@
 \end{verbatim}
-  \item Enables the construction of statically safe access operations
+  \item<2-> Enables the construction of statically safe access operations
     for vectors
@@
 \begin{verbatim}
-dotp : forall {E : Element} {{p : IsNumeric E}} {n : Nat}
-     -> PreVector n E -> PreVector n E -> Scalar E
-dotp{E} xs ys = 
+dotp : forall {n : Nat}
+     -> PreVector n Float -> PreVector n Float
+     -> Scalar Float
+dotp xs ys = 
   let xs' = use xs
@@
   in
-  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
+  fold _+_ ("0" ::: Float) (zipWith _*_ xs' ys')
 \end{verbatim}
   \begin{itemize}
-  \item \texttt{PreVector n E} : a vector of $n$ elements of
-    type \texttt{E}
-  \item \texttt{Scalar E} : code that evaluates to a scalar of type \texttt{E}
-  \item \texttt{IsNumeric E} restricts \texttt{E} to numeric types
+  \item \texttt{PreVector n Float} : a vector of $n$ elements of
+    type \texttt{Float}
+  \item \texttt{Scalar Float} : \textbf{code} that evaluates to a scalar of type \texttt{Float}
   \end{itemize}
@@
 \begin{verbatim}
-zipWith : forall {A B C} {sh} -> (Exp A -> Exp B -> Exp C)
+zipWith : forall {A B C} {sh}
+        -> (Exp A -> Exp B -> Exp C)
         -> Array sh A -> Array sh B -> Array sh C
@@
   \frametitle{Bounds Checking: fold and fold1}
-
+  \vspace{-\baselineskip}
   \begin{itemize}
@@
     reduce one dimension of an array using the operation
-  \item \texttt{fold} works for any array, but \texttt{fold1} requires
-    a non-empty array argument
+\begin{verbatim}
+fold op e []          = e
+fold op e [a]         = a
+fold op e [a1 ... an] = op (fold op e [a1 ... ai])
+                           (fold op e [ai+1 ... an])
+\end{verbatim}
+  \item \texttt{fold} works for any array, but \texttt{fold1} elides
+    the \texttt{e} argument and requires a non-empty array
   \item The latter restriction is not expressed in Haskell-Accelerate
   \end{itemize}
-  \begin{block}<2->{\vspace{-\baselineskip}}\small
+  \begin{block}<2->{\vspace{-\baselineskip}}
+\vspace{-\baselineskip}\small
 \begin{verbatim}
@@
 \end{verbatim}
+\vspace{-\baselineskip}
   \end{block}
@@
   \begin{itemize}
-  \item \texttt{fold f e} and \texttt{fold1 f}  partition their array
-    argument and apply \texttt{f} in a bracketing that is a-priori unknown
-  \item For deterministic results, \texttt{f} must be associative and
-    (in case of \texttt{fold}) \texttt{e} must be a unit of \texttt{f}
+  \item \texttt{fold op e} and \texttt{fold1 op}  partition their array
+    argument and apply \texttt{op} in a bracketing that is a-priori unknown
+  \item For deterministic results, \texttt{op} must be associative and
+    (in case of \texttt{fold}) \texttt{e} should be a unit of \texttt{op}
   \item<2-> Such restrictions can be expressed in Agda
@@
   \end{itemize}
-
 \end{block}
 \end{frame}
+
+\section{Limitation}
+
+\begin{frame}[fragile]
+  \frametitle{A Limitation}
+  \begin{itemize}
+  \item Haskell-Accelerate admits implementing a \texttt{filter}
+    operation (for vectors) with the usual meaning
+  \item<2-> Surprisingly, it cannot easily be implemented in Agda-Accelerate!
+  \item<2-> Why is that?
+  \end{itemize}
+  \begin{block}<3->{Implementation attempt}
+\begin{verbatim}
+filter : forall {n m : Nat}{E : Element}
+       -> Vector n E -> (Exp E -> Exp Bool) 
+       -> Vector m E
+\end{verbatim}
+    \begin{itemize}
+    \item Problem: What is \texttt{m}?
+    \end{itemize}
+  \end{block}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{A Workaround}
+  \begin{itemize}
+  \item Provide a special type of filtered vectors that pair each
+    element with a boolean indicating its presence
+\begin{verbatim}
+FVector : Nat -> Element -> Set
+FVector n E = Vector n (Pair Bool E)
+\end{verbatim}
+  \item<2-> Now filtering becomes straightforward 
+\begin{verbatim}
+filterF : forall {n : Nat}{E : Element}
+        -> (Exp E -> Exp Bool)
+        -> FVector n E -> FVector n E
+filterF {n}{E} pred vec =
+  map g vec
+  where
+  g : Exp (Pair Bool E) -> Exp (Pair Bool E)
+  g bx = pair ((fst bx) && p (snd bx)) x
+\end{verbatim}
+  \end{itemize}
+\end{frame}
+\begin{frame}[fragile]
+  \frametitle{More Operations on Filtered Vectors}
+  \vspace{-\baselineskip}
+  \begin{block}{Mapping}
+  \vspace{-\baselineskip}\small
+\begin{verbatim}
+mapF : forall {n : Nat}{E F : Element}
+     -> Exp F -> (Exp E -> Exp F) 
+     -> FVector n E -> FVector n F
+\end{verbatim}
+    requires a dummy element of \texttt{F} to put in the result when
+    the corresponding source entry is not present
+  \end{block}
+  \begin{block}<2->{Folding}
+  \vspace{-\baselineskip}\small
+\begin{verbatim}
+foldF : forall {n : Nat}{E : Element}
+      -> (Exp E -> Exp E -> Exp E) -> Exp E
+      -> FVector n E -> Scalar E
+foldF f e vec =
+  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
+\end{verbatim}
+    gets out of the filtered vector back to a ``real'' value
+  \end{block}
+\end{frame}
+
+\section{Wrap Up}
+
+\begin{frame}
+  \frametitle{In the paper}
+  \begin{itemize}
+  \item More information on the representation of types and constants
+  \item Further examples
+  \item Description of the implementation using the foreign function
+    interface of the Alonzo compiler
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Conclusion}
+  \begin{itemize}
+  \item Only partially successful
+  \item Bounds checking works well
+  \item Large part could be done using lifted datatypes
+  \item Generative languages do not fit well with dependent types
+  \end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Example: Polymorphic Dot Product in Agda-Accelerate}
+  \vspace{-\baselineskip}
+\begin{verbatim}
+dotp : forall {E : Element} {{p : IsNumeric E}} {n : Nat}
+     -> PreVector n E -> PreVector n E -> Scalar E
+dotp{E} xs ys = 
+  let xs' = use xs
+      ys' = use ys
+  in
+  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
+\end{verbatim}
+  \begin{itemize}
+  \item \texttt{PreVector n E} : a vector of $n$ elements of
+    type \texttt{E}
+  \item \texttt{Scalar E} : code that evaluates to a scalar of type \texttt{E}
+  \item \texttt{IsNumeric E} restricts \texttt{E} to numeric types
+  \end{itemize}
+\end{frame}
+
 \end{document}

### 2012-08-30 paper/main.tex
small fixes

@@
 \begin{code}
-dotp : {E : Element} {{p : IsNumeric E}} {n : Nat}
+dotp : forall {E : Element} {{p : IsNumeric E}} {n : Nat}
      -> PreVector n E -> PreVector n E -> Scalar E
@@
 \begin{verbatim}
-fold : {E}{sh}{n} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E)
+fold : forall {E}{sh}{n} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E)
      -> Array (sh :< n >) E -> IsMonoid f e -> Array sh E
@@
 \begin{verbatim}
-IsMonoid : {E} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E) -> Set
+IsMonoid : forall {E} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E) -> Set
 IsMonoid f e = ( IsAssociative f , IsUnit f e)
@@
 \begin{verbatim}
-filter : {n m : Nat}{E : Element}
+filter : forall {n m : Nat}{E : Element}
        -> Vector n E -> (Exp E -> Exp Bool) -> Vector m E
@@
 There are several ways of implementing this idea. The simplest
-approach is to wrap each element in a maybe type or
-pair up each element with a boolean flag that indicates
-its presence.\footnote{A \texttt{Maybe} type is on one hand the better option,
-  but it has to be coded without using pattern matching.}
+approach is to pair up each element with a boolean flag that indicates
+its presence.\footnote{The \texttt{Maybe} type is not an option
+  because Accelerate does not support \texttt{Maybe} types as array elements.}
 \begin{verbatim}
@@
 \begin{verbatim}
-filterF : {n : Nat}{E : Element}
+filterF : forall {n : Nat}{E : Element}
         -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
@@
 \begin{verbatim}
-mapF : {n : Nat}{E F : Element}
+mapF : forall {n : Nat}{E F : Element}
      -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
@@
 \begin{verbatim}
-foldF : {n : Nat}{E : Element}
+foldF : forall {n : Nat}{E : Element}
       -> (Exp E -> Exp E -> Exp E) -> Exp E
@@
 \begin{verbatim}
-use : {sh : Shape}{E : Element} -> PreArray sh E -> Array sh E
+use : forall {sh : Shape}{E : Element} -> PreArray sh E -> Array sh E
 use {sh}{E} (PA y) = Ar (useHs (EltDict E) y)
@@
 \begin{code}
-fold : {E}{sh}{n}
+fold : forall {E}{sh}{n}
      -> (Exp E -> Exp E -> Exp E)

### 2012-11-25 paper/main.tex
cosmetics, changed a few misleading formulations

@@
 \title{Agda Meets Accelerate}
-\subtitle{Extended Abstract}
+% \subtitle{Extended Abstract}
 \author{Peter Thiemann\inst{1} \and Manuel M. T. Chakravarty\inst{2}}
@@
 
-  This raises the question whether embedded languages for data parallel programming would benefit from fully-fledged dependent types, such as those available in Agda.
-We explored that question by designing and implementing an Agda frontend to Accelerate, an embedded language for data parallel programming aimed at GPUs.
+  This observation raises the question whether an embedded language
+  for data parallel programming would benefit from fully-fledged
+  dependent types, such as those available in Agda. We explored that
+  question by designing and implementing an Agda frontend to
+  Accelerate, a Haskell-embedded language for data parallel programming aimed at GPUs.
 We discuss the potential of dependent types in this domain, describe some of the limitations that we encountered, and share some insights from our preliminary implementation.
@@
 combine high-level programming models with high performance. They are
-particularly attractive for targeting restricted architectures, such
-as GPUs (graphics processor units), that cannot efficiently execute
-code aimed at conventional multicore CPUs. Instead, GPUs require a
+particularly attractive for targeting restricted architectures that cannot efficiently execute
+code aimed at conventional multicore CPUs. One prime example are GPUs
+(graphics processor units), which require a
 high degree of data parallelism, restricted control flow, and
-carefully tailored data access patterns to be efficient. Previous work
+custom tailored data access patterns to be efficient. Previous work
 ---for example, Accelerator~\cite{Tarditi:2006},
@@
 manner~\cite{ChakravartyKellerLeeMcdonellGrover2011}. This design
-choice is important for approaches relying on runtime code generation:
-compile-time faults in the embedded language should be avoided because
-it corresponds to at application runtime. Moreover, static guaranties
+choice is desirable for approaches relying on run-time code
+generation:
+each  potential fault at application run time should be discovered by
+a compile-time fault in the embedded language. Moreover, static guarantees
 hold the potential to improve the predictability of parallel
@@
 Dependent types are an emerging approach to certified programming, where invariants are established in the form of types and proven at compile time.
-Many of Haskell's type-level extensions used in Accelerate approximate various aspects of dependently-typed programming.
-Hence, it it is natural to ask whether fully-fledged dependent types, such as those provided by Agda, improve the specification of an embedded language like Accelerate, whether they increase the scope of static guarantees, and whether they may be leveraged to predict performance more accurately.
+Many of Haskell's type-level extensions used in Accelerate approximate aspects of dependently-typed programming.
+Hence, it is natural to ask whether fully-fledged dependent types, such as those provided by Agda, improve the specification of an embedded language like Accelerate, whether they increase the scope of static guarantees, and whether they may be leveraged to predict performance more accurately.
 
@@
 Section~\ref{sec:dependent-types} discusses potential uses of
-dependent types in an array-oriented data parallel language like
-Accelerate and how they were realized in our
+dependent types in an array-oriented data parallel language and how they were realized in our
 implementation. Section~\ref{sec:limitations} considers conceptual 
@@
 get []        _       ()      -- impossible case
-get (x :: xs) zero    p       = x
+get (x :: xs) zero    z<s     = x
 get (x :: xs) (suc m) (s<s p) = get xs m p
@@
 
-As computation representations of type \verb+Acc a+ are compiled at application runtime, all \verb+Acc+ compilation errors are effectivly \emph{runtime errors} of the application.
-Hence, Accelerate uses a range of Haskell type system extensions to statically type Accelerate expressions, such that these runtime errors are avoided where possible.
+As computation representations of type \verb+Acc a+ are compiled at
+application run time, all \verb+Acc+ compilation errors are effectively \emph{run-time errors} of the application.
+Hence, Accelerate uses a range of Haskell type system extensions to statically type Accelerate expressions, such that these run-time errors are avoided where possible.
 In particular, Accelerate uses GADTs\cite{PeytonJonesVytiniotisWeirichWashburn2006}, associated types
@@
 dotp :: Vector Float -> Vector Float -> Acc (Scalar Float)
-dotp xs ys = let xs' = use xs
-                 ys' = use ys
+dotp xs ys = let { xs' = use xs; ys' = use ys }
              in  fold (+) 0 (zipWith (*) xs' ys')
@@
 
-The \verb+use+ operation makes a Haskell array available in the embedded language by wrapping it into the \verb+Acc+ constructor and copying it to GPU memory.\footnote{Accelerate employs a caching strategy to avoid the transfer of arrays, which are already available in GPU memory.} The operations \texttt{fold} and \texttt{zipWith} represent collective operations on Accelerate arrays, effectively producing a representation of an array computation yielding a value of
-\texttt{Scalar Float}; i.e., a single float value. This code relies
-heavily on (type class) overloading: \texttt{0}, \texttt{(+)}, and
-\texttt{(*)} are overloaded to just construct abstract syntax.
+The \verb+use+ operation makes a Haskell array available in the
+embedded language by wrapping it into the \verb+Acc+ constructor. It
+amounts to copying it to GPU memory.\footnote{Accelerate employs a
+  caching strategy to avoid the transfer of arrays, which are already
+  available in GPU memory.} The operations \texttt{fold} and
+\texttt{zipWith} represent collective operations on Accelerate arrays,
+effectively producing a representation of an array computation
+yielding  a single float value (\texttt{Scalar Float}). The code relies on (type class) overloading: \texttt{0}, \texttt{(+)}, and
+\texttt{(*)} are overloaded to construct abstract syntax.
 
@@
 %
-where the class \verb+Elt+ characterises all types that may be held in Accelerate arrays. These are currently primitive types and tuples.
+where the class \verb+Elt+ characterizes all types that may be held in Accelerate arrays. These are currently primitive types and tuples.
 
@@
 type DIM1 = DIM0 :. Int
--- <and so on>
+-- and so on
 \end{hcode}
@@
 
-Peebles formalised parts of the API of the Haskell array library Repa using Agda~\cite{peebles:derpa}.
-The formalisation uses the same shape structure as we are using in Accelerate, but array computations are neither embedded nor can parallel high-performance code be generated.
+Peebles formalized parts of the API of the Haskell array library Repa
+using Agda~\cite{peebles:derpa}. The formalisation relies on the same
+shape structure as Accelerate, but array computations are neither
+embedded nor can parallel high-performance code be generated.
 
@@
 
-Dependent ML is an ML dialect with a restricted form of dependent types, which, among other applications, may be used to statically check array bounds in array operations~\cite{xi:dml-jfp}.
+Dependent ML is an ML dialect with a restricted form of dependent types, which, among other applications, may be used to statically check array bounds~\cite{xi:dml-jfp}.
 However, only simple indexing and array updating are considered and not aggregate array operations, such as those provided by Accelerate.
@@
 However, no attempt is made to track properties of array programs statically.
-Similarly, Copperhead~\cite{Catanzaro:EECS-2010-124} embedded an array language into Python, but does not attempt to track information statically.
+Similarly, Copperhead~\cite{Catanzaro:EECS-2010-124} embeds an array language into Python, but does not attempt to track information statically.
 
@@
 array computations. Almost all of these types have no suitable
-counterpart in Agda, which only supports natural numbers in unary
-encoding. For that reason, our embedding keeps the Haskell types
-abstract in Agda. To specify type signatures and in particular
-functions that are polymorphic in such a Haskell type or depend on it
-in some way, we have reified these types in an Agda type
-\texttt{Element}.
+counterpart in Agda, which only supplies  computationally expensive
+encodings for natural and rational numbers. For that reason, our embedding keeps the Haskell types 
+abstract in Agda. To specify the types of functions that are polymorphic in such a Haskell type or depend on it
+in some way, we have reified the possible element types in an Agda type
+\texttt{Elt}.
 \begin{verbatim}
-data Element : Set where
-  Bool   : Element
-  Int    : Element
-  Float  : Element
-  Double : Element
-  Pair   : Element -> Element -> Element
+data Elt : Set where
+  Bool   : Elt
+  Int    : Elt
+  Float  : Elt
+  Double : Elt
+  Pair   : Elt -> Elt -> Elt
   -- and so on
@@
 the set of numeric types is defined by a predicate
-\texttt{IsNumeric}.\footnote{$\top$ is a one-element type,
+\texttt{Numeric}.\footnote{$\top$ is a one-element type,
   whereas $\bot$ is a type without elements. These types customarily
@@
 \begin{alltt}
-IsNumeric : Element -> Set
-IsNumeric Int = \(\top\)
-IsNumeric Float = \(\top\)
-IsNumeric Double = \(\top\)
-IsNumeric _ = \(\bot\)
+Numeric : Elt -> Set
+Numeric Int = \(\top\)
+Numeric Float = \(\top\)
+Numeric Double = \(\top\)
+Numeric _ = \(\bot\)
 \end{alltt}
@@
 
-To see the Agda embedding in action, we translate the
+To demonstrate the Agda embedding in action, we translate the
 dot product example from Section~\ref{sec:accelerate} to Agda.
 \begin{code}
-dotp : forall {E : Element} {{p : IsNumeric E}} {n : Nat}
+dotp : forall {E : Elt} {{p : Numeric E}} {n : Nat}
      -> PreVector n E -> PreVector n E -> Scalar E
@@
       ys' = use ys
-  in
-  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
+  in  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
 \end{code}
@@
 We use them much like type class constraints are used in Haskell.} It furthermore
-takes the length $n$ as a parameter thus ensuring that the two input
+takes the length $n$ as an implicit parameter thus ensuring that the two input
 vectors have the same size. The \texttt{PreVector} type of the
@@
 \begin{code}
-use : {sh : Shape}{E : Element} -> PreArray sh E -> Array sh E
+use : {sh : Shape}{E : Elt} -> PreArray sh E -> Array sh E
 \end{code}
@@
 \begin{code}
-_+_ : {E : Element} {{p : IsNumeric E}} -> Exp E -> Exp E -> Exp E
+_+_ _*_ : {E : Elt} {{p : Numeric E}} -> Exp E -> Exp E -> Exp E
 \end{code}
@@
 abstract syntax for an addition or a multiplication by delegating to
-the corresponding Accelerate functions. The type \texttt{Exp E},
-which denotes an AST of an expression of type \texttt{E}.
+the corresponding Accelerate functions. The type \texttt{Exp E}
+denotes an AST of an expression of type \texttt{E}.
 
@@
 that no shape mismatches occur (e.g., a one-dimensional array cannot be
-considered two-dimensional). However, they do not ensure at compile
-time that the sizes of the dimensions match up.
+considered two-dimensional), but they do not ensure at compile
+time that the sizes of the dimensions match up. Such a mismatch
+results in a run-time error.
 
@@
 \begin{alltt}
-reshape : \{sh : Shape\} \{E : Element\}
+reshape : \{sh : Shape\} \{E : Elt\}
        -> (sh' : Shape) -> Array sh E -> (size sh \(\equiv\) size sh')
@@
 \end{verbatim}
-The latter type is more restrictive than the Accelerate
-implementation of \texttt{zipWith}. Instead of checking the sizes
-of the input arrays, it truncates them to the respective minima. A
-corresponding Agda type could be developed easily. It just requires a
-binary function that computes the minimum of two shapes, which is a
-simple exercise.
+The latter type is more restrictive than the Accelerate implementation
+of \texttt{zipWith}. Instead of checking the sizes of the input
+arrays, it truncates them to the respective minima. We also developed
+an Agda type that directly corresponds to this implementation. It
+requires a binary function that computes the minimum of two shapes,
+which we leave as an exercise to the reader.
 
@@
 In both cases, the text of the documentation says that ``the first
-argument needs to be associative'' and the \texttt{fold1}
+argument needs to be associative'' and furthermore the \texttt{fold1}
 documentation ``requires the reduced array to be non-empty''.
@@
 \begin{verbatim}
-fold : forall {E}{sh}{n} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E)
+fold : forall {E}{sh}{n}
+     -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E)
      -> Array (sh :< n >) E -> IsMonoid f e -> Array sh E
@@
 \begin{verbatim}
-IsMonoid : forall {E} -> (f : Exp E -> Exp E -> Exp E) -> (e : Exp E) -> Set
+IsMonoid : forall {E} -> (Exp E -> Exp E -> Exp E) -> Exp E -> Set
 IsMonoid f e = ( IsAssociative f , IsUnit f e)
 \end{verbatim}
+
+Some readers may object that neither addition nor multiplication of
+floating point numbers is associative
+\cite{DBLP:journals/csur/Goldberg91}. However, for advanced
+optimizations, the exploitation of algebraic laws is a necessity and
+the involved degradation of precision or change of result is accepted
+or account for in the error estimates. Moreover, there are
+other operations, like min or max, that are commonly used with
+\texttt{fold}-like operations, which are truly associative. Last, but
+not least, the associativity declarations serve as an important
+documentation that passing an inherently non-associative function will
+produce unpredictable, implementation-dependent results.
 
@@
 Because of Agda's limited support for numeric data types,
-we embed more ambitious numeric literals for floating point numbers
+we embed numeric literals for integers and floating point numbers
 using a string with an explicit type annotation that determines the
@@
 Recall that \texttt{Float} and \texttt{Double} are not
-types, but rather values of type \texttt{Element}.
-All magic of the embedding is hidden in the \texttt{:::} operation:
+types, but rather values of type \texttt{Elt}.
+The \texttt{:::} operation is the workhorse of the embedding:
 \begin{verbatim}
-_:::_ : (s : String) -> (E : Element) 
-     -> {{nu : IsNumeric E}} -> {p : T (s parsesAs E)} -> Exp E
+_:::_ : (s : String) -> (E : Elt) 
+     -> {{nu : Numeric E}} -> {p : T (s parsesAs E)} -> Exp E
 s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
@@
 The arguments \texttt{s} and \texttt{E} are explicit, but the
-remaining ones are to be inferred by Agda.
+remaining ones are inferred by Agda.
 As mentioned before, the argument \texttt{nu} is an instance argument; it is automatically
@@
 \cite{DevriesePiessens2011}. Here, the predicate 
-\texttt{IsNumeric} plays the role of a type class that characterizes
+\texttt{Numeric} plays the role of a type class that characterizes
 the numeric types.
@@
 The function \texttt{parsesAs} dispatches on its ``type'' argument and
-checks whether the string is a constant of the expected type. The
+parses the string to check whether it is an acceptable literal of the expected type. The
 function \texttt{constantFromString} is imported from Accelerate.
@@
 applicability of dependent typing. We already mentioned that the
-formalization of associativity or of the concept of a monoid gets
-unmanageable because such properties have to be asserted for abstract syntax.
+formalization of associativity or of the concept of a monoid cannot be
+verified in Agda because such properties have to be asserted for abstract syntax.
 
@@
 \begin{verbatim}
-filter : forall {n m : Nat}{E : Element}
+filter : forall {n m : Nat}{E : Elt}
        -> Vector n E -> (Exp E -> Exp Bool) -> Vector m E
@@
 The problem is that the size of the result
-cannot be determined statically. In fact, the only thing we know about
-\texttt{m} is that it must be less than or equal to
-\texttt{n}.
-However, we cannot prove this from the code because of a staging restriction.
+cannot be determined statically. In fact, the only thing that we could
+prove from an implementation of \texttt{filter} is that
+\texttt{m} must be less than or equal to \texttt{n}.
+However, we cannot even prove this restriction on \texttt{m} for two
+reasons. First, the code of \texttt{filter} is not available as even
+the underlying Haskell library only deals with abstract
+syntax. Second, even if the code were available, there is still a
+problem due to staging.
 The Accelerate implementation computes the length of the result only when the generated GPU code is executed.
@@
 
-We might contemplate emplying an existential type like
+We might contemplate employing an existential type like
 \begin{verbatim}
@@
 \end{verbatim}
-However, it is not possible to build such an
+but it is not possible to build such an
 existential package because the evidence \texttt{m} is not available
-when the package would have to be constructed.
+when the existential package has to be constructed.
 
@@
 approach is to pair up each element with a boolean flag that indicates
-its presence.\footnote{The \texttt{Maybe} type is not an option
+its presence.\footnote{Using an array of \texttt{Maybe}-typed elements is not an option
   because Accelerate does not support \texttt{Maybe} types as array elements.}
 \begin{verbatim}
-FVector : Nat -> Element -> Set
+FVector : Nat -> Elt -> Set
 FVector n E = Vector n (Pair Bool E)
@@
 
-Now filtering becomes quite simple because the length of
-the\texttt{FVector} does not change. Furthermore, filtering could be
+In this encoding, filtering is quite simple because the length of
+the \texttt{FVector} does not change. Furthermore, filtering could be
 extended to multi-dimensional arrays, although it is not clear if an
-array that includes non-existing elements is a sensible notion.
+array that contains non-existing elements or holes is a sensible notion.
 \begin{verbatim}
-filterF : forall {n : Nat}{E : Element}
+filterF : forall {n : Nat}{E : Elt}
         -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
@@
 \end{verbatim}
-However, mapping becomes more complicated because it either has to
+The mapping operation that applies a function to each element of an
+array becomes more complicated because it either has to 
 materialize a dummy result for each absent element in the argument
-vector or apply the function to absent elements, too.\footnote{This
-  complication could be avoided with the \texttt{Maybe} type.}
+vector or apply the function to absent elements, too.
 \begin{verbatim}
-mapF : forall {n : Nat}{E F : Element}
+mapF : forall {n : Nat}{E F : Elt}
      -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
@@
 \begin{verbatim}
-foldF : forall {n : Nat}{E : Element}
+foldF : forall {n : Nat}{E : Elt}
       -> (Exp E -> Exp E -> Exp E) -> Exp E
@@
 
-In the end, such a representation may not be a loss on a GPU. As long
+In the end, such a representation may not even lead to reduced efficiency on a GPU. As long
 as all computations take the same path, all processing elements work
@@
 computation step. So it would be most advantageous to organize work as
-uniformly as possible.
+uniformly as possible by reorganizing the array so that the present
+and the absent elements are grouped together. A segmented array might
+be a suitable representation.
 
@@
 
-We believe that these adapations only have a minor performance impact
+We believe that these adaptations only have a minor performance impact
 because (1) most functions just manipulate abstract syntax, so that
@@
 \begin{verbatim}
-use : forall {sh : Shape}{E : Element} -> PreArray sh E -> Array sh E
+use : forall {sh : Shape}{E : Elt} -> PreArray sh E -> Array sh E
 use {sh}{E} (PA y) = Ar (useHs (EltDict E) y)
@@
 Haskell via FFI) and the function \texttt{EltType} (not shown), which
-interprets a value of type \texttt{Element} as a Haskell type. The
+interprets a value of type \texttt{Elt} as a Haskell type. The
 latter types are also imported via FFI.
 \begin{code}
-data PreArray (E : Element) : Shape -> Set where
+data PreArray (E : Elt) : Shape -> Set where
   PA : {sh : Shape} -> HsArray HsDIM1 (EltType E) -> PreArray sh E
@@
 \begin{code}
-data Array (E : Element) : Shape -> Set where
+data Array (E : Elt) : Shape -> Set where
   Ar : {sh : Shape} -> Acc (AccArray HsDIM1 (EltType E)) -> Array sh E
@@
 The \texttt{EltDict} function translates a value 
-\texttt{(E : Element)} into a Haskell expression that evaluates to a
+\texttt{(E : Elt)} into a Haskell expression that evaluates to a
 dictionary for the Haskell type of \texttt{E} for the Haskell type
-class \texttt{Elt}. Such a dictionary is passed, whenever that
+class \texttt{Elt}. Such a dictionary is passed, whenever the
 corresponding Haskell function has type class constraints.
 \begin{code}
-EltDict : (E : Element) -> HsEltDict (EltType E)
+EltDict : (E : Elt) -> HsEltDict (EltType E)
 \end{code}
@@
 to include a \texttt{Shape}-indexed encoding of the \texttt{Shape}
-type in the \texttt{Element} type so that we can describe the type of
+type in the \texttt{Elt} type so that we can describe the type of
 an expression whose value has a certain shape.
@@
 Finally, the type translation of Agda's FFI has many shortcomings that
-caused a number of problems for transporting information between Agda
+cause a number of problems when transporting information between Agda
 and Haskell. One part of the problem is, unfortunately, the rich type
-structure of Accelerate which already encodes much useful
-information. An alternative, untyped (or less-typed) interface to
-Accelerate would make the adaptation to an Agda frontend much simpler.
+structure of Accelerate's API which already encodes many useful
+constraints. An alternative, untyped (or less-typed) interface to
+Accelerate would make the adaptation to an Agda frontend simpler.
 

### 2012-11-30 paper/main.tex

@@
 Haskell, which enables it to directly invoke the functionality of
-Accelerate.
-
-Our investigation has the following structure. After recalling some
+Accelerate. The main contributions of this paper are the following:
+%
+\begin{itemize}
+\item We identify and discuss the challenges of combining generative embedded languages with dependent typing (Section~\ref{sec:dependent-types}).
+\item We propose predicated arrays to overcome some of these challenges (Section~\ref{sec:limitations}).
+\item We outline an implementation of the main parts of Accelerate in Agda using the Agda-Haskell FFI for code execution (Section~\ref{sec:implementation}).
+\end{itemize}
+%
+Overall, our investigation has the following structure. After recalling some
 background on Agda and Accelerate in Section~\ref{sec:background} and
@@
 construct the proof tree for \verb+m < n+ before invoking
-\verb+get+. (In Agda, arguments in curly braces are \emph{implicit
+\verb+get+. Thus, an ``index out of bounds'' error cannot happen. 
+(In Agda, arguments in curly braces are \emph{implicit
   arguments} that will be inferred if omitted in an application.)
-Thus, an ``index out of bounds'' error cannot happen.
 
@@
 using a \verb+run+ operation that accepts such a representation (of type
-\verb+Acc a+), compiles it to GPU kernels, uploads it to the device,
+\verb+Acc a+), compiles it to GPU kernels, uploads it to a device,
 executes it, and retrieves the results.\footnote{To distinguish
@@
 Accelerate supports a wide range of numeric types, characterized by
-type class \texttt{Elt}, as base types for
-array computations. Almost all of these types have no suitable
+the type class \texttt{Elt}, as base types for
+array computations. Almost all of these types lack a suitable
 counterpart in Agda, which only supplies  computationally expensive
@@
 abstract in Agda. To specify the types of functions that are polymorphic in such a Haskell type or depend on it
-in some way, we have reified the possible element types in an Agda type
-\texttt{Elt}.
+in some way, we have reified the possible element types as an Agda type
+\texttt{Elt}:
 \begin{verbatim}
@@
 the set of numeric types is defined by a predicate
-\texttt{Numeric}.\footnote{$\top$ is a one-element type,
+\texttt{Numeric}:\footnote{$\top$ is a one-element type,
   whereas $\bot$ is a type without elements. These types customarily
@@
 approach is to pair up each element with a boolean flag that indicates
-its presence.\footnote{Using an array of \texttt{Maybe}-typed elements is not an option
+its presence, which we call \emph{predicated arrays}:\footnote{Using an array of \texttt{Maybe}-typed elements is not an option
   because Accelerate does not support \texttt{Maybe} types as array elements.}
@@
 materialize a dummy result for each absent element in the argument
-vector or apply the function to absent elements, too.
+vector or apply the function to absent elements, too. This makes \texttt{filter} reminiscent of the \texttt{where} statement of the SIMD language C$^*$~\cite{rose-etal:c-star}.
 \begin{verbatim}
@@
 
-As an example for a function that requires more work on either side,
+As an example for a function that requires more work on both sides,
 consider the \texttt{fold} operation. 

### 2012-11-30 paper/main.tex
source code ref added

@@
 example code.
+
+\noindent
+Source code is available at \url{https://github.com/mchakravarty/accelerate-agda}.
 
@@
 
-Finally, the type translation of Agda's FFI has many shortcomings that
-cause a number of problems when transporting information between Agda
+Finally, the type translation of Agda's FFI has a number of shortcomings that
+cause problems when transporting information between Agda
 and Haskell. One part of the problem is, unfortunately, the rich type

### 2013-03-04 paper/main.tex
paper revision, first round

@@
 \DefineVerbatimEnvironment{code}{Verbatim}{} % Add fancy options here if you like.
-\DefineVerbatimEnvironment{hcode}{Verbatim}{formatcom=\color{blue},frame=leftline} % Add fancy options here if you like.
+\DefineVerbatimEnvironment{hcode}{Verbatim}{formatcom=\color{blue},frame=single} % Add fancy options here if you like.
 
@@
 
-Dependent types are an emerging approach to certified programming, where invariants are established in the form of types and proven at compile time.
-Many of Haskell's type-level extensions used in Accelerate approximate aspects of dependently-typed programming.
-Hence, it is natural to ask whether fully-fledged dependent types, such as those provided by Agda, improve the specification of an embedded language like Accelerate, whether they increase the scope of static guarantees, and whether they may be leveraged to predict performance more accurately.
+Dependent types \cite{MartinLoef1984} are an established approach to certified programming,
+where invariants are established in the form of types and proven at
+compile time. Many of Haskell's type-level extensions used in
+Accelerate approximate aspects of dependently-typed
+programming. Hence, it is natural to ask whether fully-fledged
+dependent types, such as those provided by Agda, improve the
+specification of an embedded language like Accelerate, whether they
+increase the scope of static guarantees, and whether they may be
+leveraged to predict performance more accurately.
 
@@
 Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011} is a
-data-parallel array language embedded into Haskell, which targets
-GPUs. It is a \emph{generative library}, as its data-parallel array
-operations are not executed directly, but instead construct abstract
+\emph{generative} data-parallel array language embedded into Haskell, which targets
+GPUs. Being generative, its data-parallel array
+operations are not executed directly. Instead, Accelerate constructs abstract
 syntax trees (AST) representing an entire data-parallel
@@
 executes it, and retrieves the results.\footnote{To distinguish
-  Haskell code from Agda code, we display Haskell code in blue and with
-  a vertical bar on the left side.}
+  Haskell code from Agda code, we display Haskell code in a blue box.}
 %
@@
 embedded language by wrapping it into the \verb+Acc+ constructor. It
-amounts to copying it to GPU memory.\footnote{Accelerate employs a
-  caching strategy to avoid the transfer of arrays, which are already
+amounts to copying it to GPU memory.\footnote{Accelerate employs
+  caching to avoid the transfer of arrays that are already
   available in GPU memory.} The operations \texttt{fold} and
@@
 
-Peebles formalized parts of the API of the Haskell array library Repa
+Peebles formalizes parts of the Repa API % of the Haskell array library Repa
 using Agda~\cite{peebles:derpa}. The formalisation relies on the same
@@
 
-Swierstra and Altenkirch investigated the use of dependent types for \emph{distributed} array programming~\cite{swierstra-altenkirch:dep-types-for-distr-arrays,swierstra:more-dependent-types}.
-Their notation for distributed arrays was inspired by X10 and the main focus is on expressing locality aware algorithms.
+Swierstra and Altenkirch investigated the use of dependent types for
+\emph{distributed} array
+programming~\cite{swierstra-altenkirch:dep-types-for-distr-arrays,swierstra:more-dependent-types}.
+Their notation for distributed arrays is inspired by the X10 language
+\cite{Saraswat2009}. They focus on expressing locality awareness.
 
@@
 Unlike the Accelerate code, this function is polymorphic with respect
-to the array element type, provided it is numeric.\footnote{In Agda, arguments in double curly braces are \emph{instance arguments}~\cite{DevriesePiessens2011}.
-We use them much like type class constraints are used in Haskell.} It furthermore
-takes the length $n$ as an implicit parameter thus ensuring that the two input
+to the array element type, provided it is numeric.\footnote{In Agda,
+  arguments in double curly braces are \emph{instance
+    arguments}~\cite{DevriesePiessens2011} that are aggressively
+  inferred. We use them like type class constraints in Haskell.}
+The length parameter $n$ ensures that the two input
 vectors have the same size. The \texttt{PreVector} type of the
@@
 the shape of the array arguments and results. These constraints ensure
-that no shape mismatches occur (e.g., a one-dimensional array cannot be
-considered two-dimensional), but they do not ensure at compile
+that no shape mismatches occur (e.g., a 1D array cannot be
+considered 2D), but they do not ensure at compile
 time that the sizes of the dimensions match up. Such a mismatch
@@
 an Agda type that directly corresponds to this implementation. It
-requires a binary function that computes the minimum of two shapes,
+requires a binary function \texttt{isect} that computes the minimum of
+two shapes of the same rank,
 which we leave as an exercise to the reader.
-
+\begin{alltt}
+zipWith' : \{A B C\} \{shA shB\} \{p : rank shA \(\equiv\) rank shB\}
+         -> (A -> B -> C)
+         -> Array shA A -> Array shB B -> Array (isect shA shB p) C
+\end{alltt}
 \subsection{Associativity of Operations}
@@
 the \texttt{FVector} does not change. Furthermore, filtering could be
-extended to multi-dimensional arrays, although it is not clear if an
-array that contains non-existing elements or holes is a sensible notion.
+extended to multi-dimensional arrays, although the result might
+require careful interpretation.
 \begin{verbatim}
@@
         -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
-filterF {n}{E} pred vec =
-  map g vec
-  where
-  g : Exp (Pair Bool E) -> Exp (Pair Bool E)
+filterF {n}{E} pred vec = map g vec
+  where g : Exp (Pair Bool E) -> Exp (Pair Bool E)
         g bx = pair ((fst bx) && p (snd bx)) x
@@
      -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
-mapF {n}{E}{F} defaultF f vec =
-  map g vec
-  where
-  g : Exp (Pair Bool E) -> Exp (Pair Bool F)
+mapF {n}{E}{F} defaultF f vec = map g vec
+  where g : Exp (Pair Bool E) -> Exp (Pair Bool F)
         g bx = if (fst bx) then (pair (fst bx) (f (snd bx)))
@@
 \end{verbatim}
-On the positive side, some operations can get rid of the absent
-elements. In particular, a fold operation which reduces a filtered
+Some operations can get rid of absent
+elements. A fold operation which reduces a filtered
 vector with a monoid returns a single value. In Accelerate, such a value has type
@@
 \end{verbatim}
-Other operations like \texttt{fold1} and the scan operations present
-in Accelerate can also be lifted to this representation, but they
-retain a notion of absent elements and do not allow to revert to a
+Operations like \texttt{fold1} and the scan operations
+extend to this representation, but they
+% retain a notion of absent elements and
+cannot revert to a
 non-filtered representation.
@@
 
-We have build an experimental Agda frontend for the Accelerate
+We have built an experimental Agda frontend for the Accelerate
 language. The goal of this experiment was to explore potential uses of

### 2013-03-05 paper/main.tex
second round of revisions

@@
 
-The general type for \verb+use+ is
+In the general type for \verb+use+, the class \verb+Elt+ characterizes all types that may be held in Accelerate arrays. These are currently primitive types and tuples.
 %
@@
 %
-where the class \verb+Elt+ characterizes all types that may be held in Accelerate arrays. These are currently primitive types and tuples.
-
 Common dimensions, such as \verb+DIM0+, \verb+DIM1+, and so on, are
@@
 Repa~\cite{keller-etal:repa}, shapes are inductively defined using
-type-level snoc lists built from the data types \texttt{Z} and \texttt{:.}:
+type-level snoc lists built from the data types \texttt{Z} and
+\texttt{:.}. The use of snoc lists simplifies the type signatures of
+fold operations that reduce or abstract over the least significant dimensions.
 %
@@
 data sh :. i = sh :. i
-\end{hcode}
-%
-Hence, the definitions of the dimensions:
-%
-\begin{hcode}
+-- Types for often used dimensions
 type DIM0 = Z
@@
 To demonstrate the Agda embedding in action, we translate the
-dot product example from Section~\ref{sec:accelerate} to Agda.
+dot product example from Section~\ref{sec:accelerate} to Agda.\footnote{In Agda,
+  arguments in double curly braces are \emph{instance
+    arguments}~\cite{DevriesePiessens2011} that are aggressively
+  inferred. We use them like type class constraints in Haskell.}
 \begin{code}
@@
 Unlike the Accelerate code, this function is polymorphic with respect
-to the array element type, provided it is numeric.\footnote{In Agda,
-  arguments in double curly braces are \emph{instance
-    arguments}~\cite{DevriesePiessens2011} that are aggressively
-  inferred. We use them like type class constraints in Haskell.}
+to the array element type, provided it is numeric.
 The length parameter $n$ ensures that the two input
@@
 formalization is not straightforward. Moreover, even given expressions
-with a fixed meaning, there is no general shape for associative
-functions, so that proofs can only be done for special cases.
+with a fixed meaning, associativity has to be proved on a case by case
+basis.
 
@@
 the involved degradation of precision or change of result is accepted
-or account for in the error estimates. Moreover, there are
+or accounted for in the error estimates. Moreover, there are
 other operations, like min or max, that are commonly used with
 \texttt{fold}-like operations, which are truly associative. Last, but
-not least, the associativity declarations serve as an important
+not least, the associativity declarations serve as important
 documentation that passing an inherently non-associative function will
@@
 
-Because of Agda's limited support for numeric data types,
+Because of Agda's lack of support for overloaded numeric literals,
 we embed numeric literals for integers and floating point numbers
@@
 As mentioned before, the argument \texttt{nu} is an instance argument; it is automatically
-filled-in with a suitably typed value that is currently in scope
-\cite{DevriesePiessens2011}. Here, the predicate 
+filled-in with a suitably typed value in scope
+\cite{DevriesePiessens2011}. As before, the predicate 
 \texttt{Numeric} plays the role of a type class that characterizes
@@
 and \texttt{ReadDict}.
+
+This setup results in a flexible way of handling literals. It has
+worked well in our examples.
 

### 2013-03-10 paper/main.tex
Few more changes

@@
 value of type \texttt{sh} at run time, whereas the Agda
-\texttt{reshape} requires a \texttt{Shape} as its first argument.
+\texttt{reshape} requires a \texttt{Shape} as its first argument. Hence, 
+Agda \texttt{reshape} computes the shape in the host language on the CPU,
+whereas the original signature admits to compute the new shape in the
+embedded language as part of a GPU computation. In other words, we slightly
+restrict expressiveness here to gain more static information, we will get
+back to that issue when discussing filtering.
 
-Furthermore, functions like \texttt{map} and \texttt{zipWith} obtain
+In Agda, functions like \texttt{map} and \texttt{zipWith} obtain
 more precise types. The type of \texttt{map} tells us that the input
@@
 remaining ones are inferred by Agda.
-As mentioned before, the argument \texttt{nu} is an instance argument; it is automatically
+As mentioned, the argument \texttt{nu} is an instance argument; it is automatically
 filled-in with a suitably typed value in scope
@@
 and \texttt{ReadDict}. 
-
-This setup results in a flexible way of handling literals. It has
+This results in a flexible way of handling literals, which
 worked well in our examples.
@@
 \end{verbatim}
-The problem is that the size of the result
-cannot be determined statically. In fact, the only thing that we could
-prove from an implementation of \texttt{filter} is that
-\texttt{m} must be less than or equal to \texttt{n}.
-However, we cannot even prove this restriction on \texttt{m} for two
-reasons. First, the code of \texttt{filter} is not available as even
-the underlying Haskell library only deals with abstract
-syntax. Second, even if the code were available, there is still a
-problem due to staging.
-The Accelerate implementation computes the length of the result only when the generated GPU code is executed.
-The function \texttt{Exp E -> Exp Bool} maps abstract syntax to abstract syntax; it does not directly implement a Boolean predicate.
-Hence, we cannot use the predicate in the result type of \texttt{filter} to more accurately constrain the size of the resulting vector --- unless we include an evaluator for Accelerate expressions.
+The problem is that the size of the result cannot be determined statically --- that is,
+we cannot simplify define a type-level function that determines the length of a vector.
+Why? \texttt{Vector m E} is not a representation of a vector. Instead, it is a representation
+of a computation that, \emph{once run,} produces a vector.
+
+Similarly, we cannot define a function that uses the predicate passed to \texttt{filter} to 
+count the number of elements that will appear in \texttt{filter}'s result. Such a function would
+need access to the elements of the \texttt{filter}ed vector, but, as discussed, we cannot even get
+its length. Moreover, such a counting function would need to evaluate the predicate. We cannot do
+that as the predicate of type \texttt{Exp E -> Exp Bool} maps abstract syntax to abstract syntax;
+it \emph{does not} directly implement a Boolean predicate. We might consider to include an evaluator
+for abstract syntax to lift these restrictions. However, that evaluator would not be the code
+actually executed on the GPU, and hence, it doesn't seem to be any more valuable than simply
+asserting an axiom concerning the size of \texttt{filter}'s result.
+That is the price we pay for a generative approach, where at program runtime, we dynamically
+generate the code to be executed on the GPU.
+
+We encounter similar restrictions if we try to, at least, establish that \texttt{m} must be less
+than or equal to \texttt{n} for \texttt{filter}. We cannot prove this constraint as the GPU code of
+\texttt{filter} is not available to us --- it is generated by the underlying Haskell library. Even
+if we had access to that code, any statements about its properties would need to be based on the
+semantics of CUDA (i.e., NVIDIA's C dialect for GPU programming).
 
@@
 
-However, an alternative encoding of arrays can be
-used which is compatible with filtering of elements. The idea of this
-encoding is to keep all elements but mark those which are no longer
+However, we may use an alternative encoding of arrays that is compatible with filtering. The idea
+is to keep all elements but mark those which are no longer
 present because they have been filtered out.
@@
 approach is to pair up each element with a boolean flag that indicates
-its presence, which we call \emph{predicated arrays}:\footnote{Using an array of \texttt{Maybe}-typed elements is not an option
-  because Accelerate does not support \texttt{Maybe} types as array elements.}
+its presence, which we call \emph{predicated arrays}:\footnote{Accelerate currently does not support
+\texttt{Maybe} types as array elements.}
+%
 \begin{verbatim}
@@
 \end{verbatim}
-
+%
 In this encoding, filtering is quite simple because the length of
@@
 \end{verbatim}
-The mapping operation that applies a function to each element of an
-array becomes more complicated because it either has to 
+Mapping, which applies a function to each element of an
+array, becomes more complicated as it either has to 
 materialize a dummy result for each absent element in the argument
@@
 Alternatively, an interactively developed program may be compiled to
-Haskell using the Alonzo compiler. This compiler supports a Haskell
-foreign function interface (FFI), which enables Agda programs to invoke
+Haskell using the Alonzo compiler. It supports a Haskell
+foreign function interface (FFI), for Agda programs to invoke
 Haskell functions. 
-
 Using this interface amounts to declaring a typed identifier in Agda
@@
 one mentioned in Section~\ref{sec:array-types}. This difference arises
-because the type translation of Alonzo is unable to cope with the
-index type \texttt{Shape}. For that reason, the interface uses a
+as the type translation of Alonzo is unable to cope with the
+index type \texttt{Shape}. Hence, the interface uses a
 simplified array type and adapter functions are required, in the worst
